---
title: "Cross-isthmian denticle assemblage composition"
format: html
editor: visual
---

## Overview

Here, we characterize denticle assemblage structure to make comparisons between time periods and regions. This script cleans and filters the raw denticle count data (counts of each functional morphotype), runs the analysis, and plots Figure 2 and the related supplementary figures.

Author: Erin Dillon, emdillon23\@gmail.com

## Libraries

```{r}
#| warning: false

library(vegan)
library(permute)
library(glmmTMB)
library(DHARMa)
library(effects)
library(emmeans)
library(psych)
library(plotrix)
library(tidyverse)
library(patchwork)

```

## Data

### Read in raw data

```{r}

#Counts of each denticle functional morphotype
dd_func_pan_raw <- read.csv("data/cross_isthmus_dd_func.csv",header=T,na.strings=".",stringsAsFactors=TRUE)

#Cleaned denticle count information, for merging
dd_pan_age_clean_pre <- read.csv("clean/dd_pan_age_clean_pre.csv",header=T,na.strings="NA") %>% 
  select(-c(X))

```

### Data cleaning

Merge with relevant metadata from denticle count dataframe.

```{r}

dd_func_pan_age <- dd_func_pan_raw %>%
  #add sed_weight_kg and no_years
  left_join(dd_pan_age_clean_pre[,c(6,9,33)],by="unique") %>% 
  #calculating n demersal ABR and RA denticles
  mutate(abr_dem = abr_tot - abr_pel, 
         rabr_dem = rabr_tot - rabr_pel) %>% 
  mutate(age_group = fct_relevel(age_group,c("pre_exp","modern")))%>%
  #keeping just the relevant samples for each time period
  filter(age_group %in% c("modern","pre_exp")) %>%
  #filtering out one site with difficult to interpret time-averaging
  filter(site_name != "Airport Point") %>%
  #create grouping variable
  unite(sample_group, c("sample","group"),remove=F,sep="_") 

```

Group samples with low N to create more representative samples.

```{r}

dd_func_pan <- dd_func_pan_age %>% 
  #group together samples with low N
  group_by(basin,region,site_name,sample,sample_group,age_group) %>% 
  #summarize counts per group
  summarize(abr_grouped = sum(abr_tot,na.rm=T),
            defense_grouped = sum(defense,na.rm=T),
            drag_grouped = sum(drag,na.rm=T), 
            gen_func_grouped = sum(gen_func,na.rm=T),
            rabr_grouped = sum(rabr_tot,na.rm=T),
            abr_pel_grouped = sum(abr_pel,na.rm=T),
            rabr_pel_grouped = sum(rabr_pel,na.rm=T),
            abr_dem_grouped = sum(abr_dem,na.rm=T),
            rabr_dem_grouped = sum(rabr_dem,na.rm=T),
            dd_count_grouped = sum(total_dd_count,na.rm=T)) %>% 
  #denticle counts from pelagic & demersal sharks
  mutate(pelagic_count = drag_grouped + abr_pel_grouped + rabr_pel_grouped,
         demersal_count = abr_dem_grouped + rabr_dem_grouped,
         pel_dem_ratio = pelagic_count/demersal_count,
         prop_pelagic = pelagic_count/dd_count_grouped,
         prop_demersal = demersal_count/dd_count_grouped,
  #percent abundance of each functional morphotype
         abr_perc=(abr_grouped/dd_count_grouped)*100,
         defense_perc=(defense_grouped/dd_count_grouped)*100,
         drag_perc=(drag_grouped/dd_count_grouped)*100,
         gen_func_perc=(gen_func_grouped/dd_count_grouped)*100,
         rabr_perc=(rabr_grouped/dd_count_grouped)*100,
         rabr_pel_perc=(rabr_pel_grouped/dd_count_grouped)*100,
         abr_pel_perc=(abr_pel_grouped/dd_count_grouped)*100,
         rabr_dem_perc=(rabr_dem_grouped/dd_count_grouped)*100,
         abr_dem_perc=(abr_dem_grouped/dd_count_grouped)*100) %>% 
  ungroup()


#Create combined variable group for plotting
dd_func_pan <- dd_func_pan %>%
  unite(age_group_region, c("basin","age_group"),remove=F) %>% 
  mutate(age_group_region=as.factor(age_group_region))

```

### Sample sizes for analysis

```{r}

#samples by basin, after cleaning and filtering
table(dd_func_pan$basin)

#samples by basin, before grouping
table(dd_func_pan_age$basin)

#total count classified denticles
sum(dd_func_pan$dd_count_grouped,na.rm=T)

#overall mean denticle count per grouped sample
mean(dd_func_pan$dd_count_grouped)

#denticle count per basin
dd_func_pan %>% 
  group_by(basin) %>% 
  summarize(dd = sum(dd_count_grouped,na.rm=T))

#denticle count by basin and time period
dd_func_pan %>% 
  group_by(basin,age_group) %>% 
  summarize(dd = sum(dd_count_grouped,na.rm=T),
            n = n(),
            mean_per_sample = mean(dd_count_grouped))

#sample sizes by basin, site, and time period for table S2
dd_func_pan_age %>% 
  group_by(basin,age_group,site_name) %>% 
  summarize(dd = sum(total_dd_count,na.rm=T),
            n = n())

#sampling effort per grouped sample
dd_func_pan_age %>% 
  group_by(basin, age_group, sample_group) %>% 
  summarize(sed_weight_kg_group = sum(sed_weight_kg,na.rm=T),
            no_years_group_median = median(no_years,na.rm=T),
            dd_count_group = sum(total_dd_count,na.rm=T))

```

## Compositional analysis (Figure 2)

### Relative abundance of pelagic denticles (Fig. 2A)

```{r}

fig2A <- ggplot(dd_func_pan,aes(x=prop_pelagic*100))+
  geom_vline(xintercept=50,color="gray70",linewidth=0.6)+
  geom_density(alpha=0.6, position="identity",aes(color=region,fill=age_group_region),linewidth=1)+
  theme_bw()+
  labs(x="Percent pelagic denticles (%)",y="Density",color="Age group",linetype="Age group") +
  scale_color_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_fill_manual(values=c("darkgoldenrod2","gray80","deepskyblue3","gray80"),breaks=c("Caribbean_pre_exp","Caribbean_modern" ,"Pacific_pre_exp","Pacific_modern"),labels=c("Caribbean_pre_exp","Caribbean_modern" ,"Pacific_pre_exp","Pacific_modern"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  theme(panel.grid.major.y = element_blank())+
  xlim(0,100)+
  theme(legend.position = "none")+
  theme(aspect.ratio=0.5)+
  #guides(color="none")+
  facet_grid(~region)+
  theme(strip.background = element_rect(fill = NA, color = NA),
        strip.text = element_text(size = 16))
fig2A

```

Save output (Fig. 2A).

```{r}

#ggsave("figures/fig2A.pdf",plot=fig2A,device="pdf",useDingbats=FALSE)

```

Calculate group means for both pelagic and demersal types.

```{r}

#Pacific
mean(dd_func_pan$prop_pelagic[dd_func_pan$age_group_region=="Pacific_pre_exp"])
mean(dd_func_pan$prop_pelagic[dd_func_pan$age_group_region=="Pacific_modern"])

mean(dd_func_pan$prop_demersal[dd_func_pan$age_group_region=="Pacific_pre_exp"])
mean(dd_func_pan$prop_demersal[dd_func_pan$age_group_region=="Pacific_modern"])

#Caribbean
mean(dd_func_pan$prop_pelagic[dd_func_pan$age_group_region=="Caribbean_pre_exp"])
mean(dd_func_pan$prop_pelagic[dd_func_pan$age_group_region=="Caribbean_modern"])

mean(dd_func_pan$prop_demersal[dd_func_pan$age_group_region=="Caribbean_pre_exp"])
mean(dd_func_pan$prop_demersal[dd_func_pan$age_group_region=="Caribbean_modern"])

```

### Relative abundance of each functional morphotype (Fig. 2B)

Prepare data for plotting.

```{r}

#Create a longform dataframe
perc_long_disagg <- dd_func_pan %>% 
  select(basin,region,site_name,sample,sample_group,age_group,abr_pel_perc,abr_dem_perc,defense_perc,drag_perc,gen_func_perc,rabr_pel_perc,rabr_dem_perc) %>%
  rename(drag=drag_perc,
         abr_pel=abr_pel_perc,
         abr_dem=abr_dem_perc,
         rdgabr_pel=rabr_pel_perc,
         rdgabr_dem=rabr_dem_perc,
         gf=gen_func_perc,
         defense=defense_perc) %>% 
  gather(drag,abr_pel,abr_dem,rdgabr_pel,rdgabr_dem,gf,defense,key="func_morph",value="perc_abund") %>%
  mutate(func_morph = as_factor(func_morph)) %>% 
  arrange(sample) 


#Summarize percent abundances per region and time period for plotting
summary_comb_disagg <- perc_long_disagg %>% 
  group_by(basin,age_group,func_morph) %>% 
  summarize(abund_mean=mean(perc_abund, na.rm=T), 
            abund_sd=sd(perc_abund, na.rm=T),
            abund_stderr=std.error(perc_abund, na.rm=T)) %>% 
  rename(morph=func_morph) %>%
  #variable for plotting
  unite(age_group_region, c("basin","age_group"),remove=F) %>% 
  mutate(age_group_region=as.factor(age_group_region))

summary_comb_disagg

```

Create diverging bar plots.

```{r fig.width=4}

#Caribbean
carib_ranks <- summary_comb_disagg %>% 
  filter(basin=="Caribbean") %>% 
ggplot(aes(y=morph, x=abund_mean,fill=age_group_region,color=basin))+
  geom_bar(aes(color=basin),stat="identity",position="dodge",linewidth=0.8,alpha=0.7)+
  geom_errorbarh(aes(xmin=abund_mean-abund_stderr, xmax=abund_mean+abund_stderr), height=.2,position=position_dodge(.9),color="gray30",linetype=1,size=0.6)+
  theme_bw()+
  scale_y_discrete(limits=c("defense","gf","abr_dem","rdgabr_dem","abr_pel","rdgabr_pel","drag"),labels=c("Defense (7)","Generalized functions (6)","Abrasion strength\nDemersal (5)","Ridged abrasion strength\nDemersal (4)","Abrasion strength\nPelagic (3)","Ridged abrasion strength\nPelagic (2)","Drag reduction (1)"))+
  scale_color_manual(values=c("darkgoldenrod2"))+
  scale_fill_manual(values=c("darkgoldenrod2","gray80"),labels=c("Caribbean_pre_exp","Caribbean_modern" ,"Pacific_pre_exp","Pacific_modern"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans"))+
  theme(axis.title.x = element_text(vjust=-0.4))+
  theme(legend.position="none")+
  labs(y="Functional morphotype", x="Percent abundance (%)",fill="Age group") +
  coord_cartesian(xlim = c(60, 0))+
  theme(aspect.ratio=1.2)
carib_ranks


#Pacific
pac_ranks <- summary_comb_disagg %>% 
  filter(basin=="Pacific") %>% 
ggplot(aes(y=morph, x=abund_mean,fill=age_group_region,color=basin))+
  geom_bar(aes(color=basin),stat="identity",position="dodge",linewidth=0.8,alpha=0.7)+
  geom_errorbarh(aes(xmin=abund_mean-abund_stderr, xmax=abund_mean+abund_stderr), height=.2,position=position_dodge(.9),color="gray30",linetype=1,size=0.6)+
  theme_bw()+
  scale_y_discrete(limits=c("defense","gf","abr_dem","rdgabr_dem","abr_pel","rdgabr_pel","drag"),labels=c("Defense (7)","Generalized Functions (6)","Abrasion Strength\nDemersal (5)","Ridged Abrasion Strength\nDemersal (4)","Abrasion Strength\nPelagic (3)","Ridged Abrasion Strength\nPelagic (2)","Drag Reduction (1)"))+
  scale_color_manual(values=c("deepskyblue3"))+
  scale_fill_manual(values=c("deepskyblue3","gray80"),labels=c("Caribbean_pre_exp","Caribbean_modern" ,"Pacific_pre_exp","Pacific_modern"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans"))+
  theme(axis.title.x = element_text(vjust=-0.4))+
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())+
  theme(legend.position="none")+
  labs(y="Functional morphotype", x="Percent abundance (%)",fill="Age group") +
  coord_cartesian(xlim = c(0, 60))+
  theme(aspect.ratio=1.2)
pac_ranks

```

Combine and save output (Fig. 2B)

```{r fig.width=4}

fig2B <- carib_ranks+pac_ranks
fig2B

#save output
#ggsave("figures/fig2B.pdf",plot=fig2B,device="pdf",useDingbats=FALSE)

```

### Relative abundance by site (figure S12)

```{r}

#Summarize relative abundances at site level
summary_site_disagg <- perc_long_disagg %>% 
  group_by(basin,site_name,age_group,func_morph) %>% 
  summarize(abund_mean=mean(perc_abund, na.rm=T), 
            abund_sd=sd(perc_abund, na.rm=T),
            abund_stderr=std.error(perc_abund, na.rm=T)) %>% 
  rename(morph=func_morph)
summary_site_disagg

#Calculate sample sizes, for plotting
sample_sizes_func <- dd_func_pan %>%
  group_by(basin,site_name,age_group) %>%
  summarize(tot=sum(dd_count_grouped,na.rm=T)) %>% 
  unite(site_age, c("site_name","age_group"),remove=F) %>%
  mutate(site_age=factor(site_age, levels = c("Sweet Bocas_pre_exp","Casa Blanca_modern", "Cayo Adriana_modern","Punta Donato_modern","Contadora_pre_exp","Contadora_modern","Isla Iguana_pre_exp","Isla Iguana_modern","Saboga_pre_exp","Saboga_modern"),labels=c("Sweet Bocas (pre-exploitation)","Casa Blanca (recent)","Cayo Adriana (recent)","Punta Donato (recent)","Contadora (pre-exploitation)","Contadora (recent)","Isla Iguana (pre-exploitation)","Isla Iguana (recent)","Saboga (pre-exploitation)","Saboga (recent)"))) %>% 
  mutate(label=paste0("N=",tot)) %>% 
  mutate(x=60,y=1)

#Create variable for plotting
summary_site_disagg <- summary_site_disagg %>%
  unite(site_age, c("site_name","age_group"),remove=F) %>% 
  mutate(site_age=factor(site_age, levels = c("Sweet Bocas_pre_exp","Casa Blanca_modern", "Cayo Adriana_modern","Punta Donato_modern","Contadora_pre_exp","Contadora_modern","Isla Iguana_pre_exp","Isla Iguana_modern","Saboga_pre_exp","Saboga_modern"),labels=c("Sweet Bocas (pre-exploitation)","Casa Blanca (recent)","Cayo Adriana (recent)","Punta Donato (recent)","Contadora (pre-exploitation)","Contadora (recent)","Isla Iguana (pre-exploitation)","Isla Iguana (recent)","Saboga (pre-exploitation)","Saboga (recent)")))

```

Plot.

```{r}

figS12 <- ggplot(summary_site_disagg, aes(y=morph, x=abund_mean,fill=basin))+
  geom_bar(stat="identity",position="dodge",linewidth=1,alpha=0.8)+
  geom_errorbarh(aes(xmin=abund_mean-abund_stderr, xmax=abund_mean+abund_stderr), height=.2,
                 position=position_dodge(.9),color="gray30",linetype=1,size=0.6)+
  theme_bw()+
  scale_y_discrete(limits=c("defense","gf","abr_dem","rdgabr_dem","abr_pel","rdgabr_pel","drag"),labels=c("Defense","Generalized functions","Abrasion strength (Demersal)","Ridged abrasion strength (Demersal)","Abrasion strength (Pelagic)","Ridged abrasion strength (Pelagic)","Drag reduction"))+
  geom_text(data = sample_sizes_func, aes(x = x, y = y, label = label,inherit.aes = FALSE), size = 3,fontface = "italic")+
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans"))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.title.x = element_text(vjust=-0.4))+
  theme(legend.position="bottom")+
  labs(y="Functional morphotype", x="Percent abundance (%)",fill="Region") +
  coord_cartesian(xlim = c(0, 70))+
  theme(aspect.ratio=1)+
  facet_wrap(~site_age)+
  theme(strip.background = element_rect(fill = NA, color = NA),
        strip.text = element_text(size = 8.5))
figS12

```

Save output (fig. S11).

```{r}

#ggsave("figures/figS12.png",plot=figS12,device="png",width=10)

```

### NMDS ordination (Fig. 2C)

Prepare data for plotting and compositional analysis.

```{r}

#Make wideform dataframe with counts
dd_pan_wide_disagg <- dd_func_pan %>% 
  #variables for plotting
  unite(age_group_region, c("basin","age_group"),remove=F) %>% 
  mutate(age_group_region=as.factor(age_group_region)) %>%
  unite(age_group_site, c("site_name","age_group"),remove=F) %>% 
  mutate(age_group_site=as.factor(age_group_site)) %>%
  #keep identifiable morphotypes only (no "unknown")
  dplyr::select(basin, region, site_name, sample,sample_group, age_group, age_group_region, age_group_site, drag_grouped, abr_dem_grouped, abr_pel_grouped, rabr_dem_grouped, rabr_pel_grouped, gen_func_grouped, defense_grouped)  

```

Run NMDS.

```{r}

#Run nmds, using horn index
func_nmds_comp_disagg <- metaMDS(dd_pan_wide_disagg[,9:15],distance="horn",k=3,trymax=50)

#Check stress
stressplot(func_nmds_comp_disagg)
func_nmds_comp_disagg$stress

#Extract NMDS output & bind with metadata
nmds_df_comp_disagg <- as_tibble(func_nmds_comp_disagg$points) %>% 
  bind_cols(dd_pan_wide_disagg[,1:8],.)

```

Calculate and extract age-region ellipses for ggplot figure.

```{r}

#Extract loadings
NMDS_disagg = data.frame(MDS1 = func_nmds_comp_disagg$points[,1], MDS2 = func_nmds_comp_disagg$points[,2],group=dd_pan_wide_disagg$age_group_region)

#1SD confidence ellipses
plot(func_nmds_comp_disagg)
ord_disagg <- ordiellipse(func_nmds_comp_disagg, nmds_df_comp_disagg$age_group_region, display = "sites", kind = "sd")


#Function to extract ellipse data
veganCovEllipse <- function (cov, center = c(0, 0), scale = 1, npoints = 100) 
  {
    theta <- (0:npoints) * 2 * pi/npoints
    Circle <- cbind(cos(theta), sin(theta))
    t(center + scale * t(Circle %*% chol(cov)))
}



#Add ellipse data to df with NMDS loadings
df_combined_disagg_horn <- data.frame()
for(g in levels(NMDS_disagg$group)){
  df_combined_disagg_horn <- rbind(df_combined_disagg_horn, cbind(as.data.frame(with(NMDS_disagg[NMDS_disagg$group==g,],
                  veganCovEllipse(ord_disagg[[g]]$cov,ord_disagg[[g]]$center,ord_disagg[[g]]$scale)))
                                ,group=g))
}

#Add columns for region and time period
df_combined_disagg_horn <- df_combined_disagg_horn %>% 
  mutate(region = case_when(
    group=="Caribbean_modern" ~ "Caribbean",
    group=="Caribbean_pre_exp" ~ "Caribbean",
    group=="Pacific_modern" ~ "Pacific",
    group=="Pacific_pre_exp" ~ "Pacific")) %>% 
  mutate(age_group = case_when(
    group=="Caribbean_modern" ~ "Recent",
    group=="Caribbean_pre_exp" ~ "Pre-exploitation",
    group=="Pacific_modern" ~ "Recent",
    group=="Pacific_pre_exp" ~ "Pre-exploitation"))

```

Calculate vectors.

```{r}

env_func_disagg <- envfit(func_nmds_comp_disagg, dd_pan_wide_disagg[,9:15], permu = 99999,display='sites')
env_func_disagg

spp_scrs_disagg <- as.data.frame(scores(env_func_disagg, display = "vectors"))
spp_scrs_disagg <- cbind(spp_scrs_disagg, Morph = rownames(spp_scrs_disagg))

```

Create formatted NMDS for main figure.

```{r}

fig2C <- ggplot(nmds_df_comp_disagg, aes(x = MDS1, y = MDS2)) +
  geom_hline(yintercept=0, color="gray60")+
  geom_vline(xintercept=0, color="gray60")+
  geom_point(aes(shape=age_group_region,color=basin),size = 4, alpha = 0.6)+
  geom_polygon(data=df_combined_disagg_horn, aes(x=NMDS1, y=NMDS2,color=region,fill=group), size=1,alpha=0.4)+
   geom_segment(data = spp_scrs_disagg,
                aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
                size=0.8,
                colour = "gray30",
                arrow=arrow(length=unit(0.03,"npc")),
                linejoin='mitre') +
  geom_text(data = spp_scrs_disagg, aes(x = NMDS1*1.15, y = NMDS2*1.15, label = Morph), size = 3)+
  theme_bw()+
  scale_color_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_fill_manual(values=c("darkgoldenrod2","gray80","deepskyblue3","gray80"),breaks=c("Caribbean_pre_exp","Caribbean_modern" ,"Pacific_pre_exp","Pacific_modern"),labels=c("Caribbean_pre_exp","Caribbean_modern" ,"Pacific_pre_exp","Pacific_modern"))+
  scale_shape_manual(values=c(16,17,16,17),breaks=c("Caribbean_pre_exp","Caribbean_modern" ,"Pacific_pre_exp","Pacific_modern"),labels=c("Caribbean_pre_exp","Caribbean_modern" ,"Pacific_pre_exp","Pacific_modern"))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  xlab("NMDS 1")+
  ylab("NMDS 2")+
  labs(color="Region",shape="Age group",fill="Age group")+
  guides(color="none")+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans"))+
  theme(legend.position="right")+
  theme(aspect.ratio=1)
fig2C

```

Save output (Fig. 2C).

```{r fig.width=4}

#ggsave("figures/fig2C.pdf",plot=fig2C,device="pdf",useDingbats=FALSE)

```

### NMDS run separately by region (figure S13)

Caribbean

```{r}

#Wideform dataframe for the Caribbean
dd_carib_wide <- dd_func_pan %>% 
  filter(basin=="Caribbean") %>% 
  unite(age_group_region, c("basin","age_group"),remove=F) %>% 
  mutate(age_group_region=as.factor(age_group_region)) %>%
  unite(age_group_site, c("site_name","age_group"),remove=F) %>% 
  mutate(age_group_site=as.factor(age_group_site)) %>%
  dplyr::select(basin, region, site_name, sample,sample_group, age_group, age_group_region, age_group_site, drag_grouped, abr_dem_grouped, abr_pel_grouped, rabr_dem_grouped, rabr_pel_grouped, gen_func_grouped, defense_grouped) 


#Run NMDS with Horn index
nmds_carib <- metaMDS(dd_carib_wide[,9:15],distance="horn",k=3,trymax=50)
stressplot(nmds_carib)
nmds_carib$stress

#Extract NMDS outputs
nmds_carib_df <- as_tibble(nmds_carib$points) %>% 
  bind_cols(dd_carib_wide[,1:8],.) 

#Calculate 1SD ellipses for age groups
NMDS_carib = data.frame(MDS1 = nmds_carib$points[,1], MDS2 = nmds_carib$points[,2],group=dd_carib_wide$age_group)

plot(nmds_carib)
ord_carib <- ordiellipse(nmds_carib, nmds_carib_df$age_group, display = "sites", kind = "sd")

#Add to dataframe with NMDS loadings
df_carib_horn <- data.frame()
for(g in levels(NMDS_carib$group)){
  df_carib_horn <- rbind(df_carib_horn, cbind(as.data.frame(with(NMDS_carib[NMDS_carib$group==g,],
                  veganCovEllipse(ord_carib[[g]]$cov,ord_carib[[g]]$center,ord_carib[[g]]$scale)))
                                ,group=g))
}

#Add column for age group
df_carib_horn <- df_carib_horn %>%
  mutate(age_group = case_when(
    group=="modern" ~ "Recent",
    group=="pre_exp" ~ "Pre-exploitation"))

#Relevel
levels(nmds_carib_df$age_group) <- c("Pre-exploitation","Recent")


#Calculate vectors
env_carib <- envfit(nmds_carib,dd_carib_wide[,9:15], permu = 99999,display='sites')
env_carib

spp_scrs_carib <- as.data.frame(scores(env_carib, display = "vectors"))
spp_scrs_carib <- cbind(spp_scrs_carib, Morph = rownames(spp_scrs_carib))

```

```{r}

#Plot
figS13A <- ggplot(nmds_carib_df, aes(x = MDS1, y = MDS2)) +
  geom_hline(yintercept=0, color="gray60")+
  geom_vline(xintercept=0, color="gray60")+
  geom_point(aes(shape=age_group),color="darkgoldenrod2",size = 4, alpha = 0.6)+
  geom_polygon(data=df_carib_horn, aes(x=NMDS1, y=NMDS2,fill=age_group),color="darkgoldenrod2", size=1,alpha=0.4)+
  geom_segment(data = spp_scrs_carib,
               aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               size=0.8,
               colour = "gray30",
               arrow=arrow(length=unit(0.03,"npc")),
               linejoin='mitre') +
  geom_text(data = spp_scrs_carib, aes(x = NMDS1*1.15, y = NMDS2*1.15, label = Morph), size = 3)+
  scale_fill_manual(values=c("darkgoldenrod2","gray80"))+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  xlab("NMDS 1")+
  ylab("NMDS 2")+
  labs(shape="Time period",fill="Time period")+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans"))+
  theme(legend.position="bottom")+
  theme(aspect.ratio=1)
figS13A

```

Pacific

```{r}

#Wideform dataframe for the Pacific
dd_pac_wide <- dd_func_pan %>% 
  filter(basin=="Pacific") %>% 
  unite(age_group_region, c("basin","age_group"),remove=F) %>% 
  mutate(age_group_region=as.factor(age_group_region)) %>%
  unite(age_group_site, c("site_name","age_group"),remove=F) %>% 
  mutate(age_group_site=as.factor(age_group_site)) %>%
  dplyr::select(basin, region, site_name, sample,sample_group, age_group, age_group_region, age_group_site, drag_grouped, abr_dem_grouped, abr_pel_grouped, rabr_dem_grouped, rabr_pel_grouped, gen_func_grouped, defense_grouped) 


#Run NMDS with Horn index
nmds_pac <- metaMDS(dd_pac_wide[,9:15],distance="horn",k=3,trymax=50)
stressplot(nmds_pac)
nmds_pac$stress

#Extract NMDS outputs
nmds_pac_df <- as_tibble(nmds_pac$points) %>% 
  bind_cols(dd_pac_wide[,1:8],.)

#Calculate 1SD ellipses for age groups
NMDS_pac = data.frame(MDS1 = nmds_pac$points[,1], MDS2 = nmds_pac$points[,2],group=dd_pac_wide$age_group)

plot(nmds_pac)
ord_pac <- ordiellipse(nmds_pac, nmds_pac_df$age_group, display = "sites", kind = "sd")

#Add to dataframe with NMDS loadings
df_pac_horn <- data.frame()
for(g in levels(NMDS_pac$group)){
  df_pac_horn <- rbind(df_pac_horn, cbind(as.data.frame(with(NMDS_pac[NMDS_pac$group==g,],
                  veganCovEllipse(ord_pac[[g]]$cov,ord_pac[[g]]$center,ord_pac[[g]]$scale)))
                                ,group=g))
}

#Add column for age group
df_pac_horn <- df_pac_horn %>% 
  mutate(age_group = case_when(
    group=="modern" ~ "Recent",
    group=="pre_exp" ~ "Pre-exploitation"))

#Relevel
levels(nmds_pac_df$age_group) <- c("Pre-exploitation","Recent")


#Calculate vectors
env_pac <- envfit(nmds_pac, dd_pac_wide[,9:15], permu = 99999,display='sites')
env_pac

spp_scrs_pac <- as.data.frame(scores(env_pac, display = "vectors"))
spp_scrs_pac <- cbind(spp_scrs_pac, Morph = rownames(spp_scrs_pac))

```

```{r}

#Plot
figS13B <- ggplot(nmds_pac_df, aes(x = MDS1, y = MDS2)) +
  geom_hline(yintercept=0, color="gray60")+
  geom_vline(xintercept=0, color="gray60")+
  geom_point(aes(shape=age_group),color="deepskyblue3",size = 4, alpha = 0.6)+
  geom_polygon(data=df_pac_horn, aes(x=NMDS1, y=NMDS2,fill=age_group),color="deepskyblue3", size=1,alpha=0.4)+
  geom_segment(data = spp_scrs_pac,
               aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               size=0.8,
               colour = "gray30",
               arrow=arrow(length=unit(0.03,"npc")),
               linejoin='mitre') +
  geom_text(data = spp_scrs_pac, aes(x = NMDS1*1.15, y = NMDS2*1.15, label = Morph), size = 3)+
  scale_fill_manual(values=c("deepskyblue3","gray80"))+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  xlab("NMDS 1")+
  ylab("NMDS 2")+
  labs(shape="Time period",fill="Time period")+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans"))+
  theme(legend.position="bottom")+
  theme(aspect.ratio=1)
figS13B

```

Combine and save output (fig S13).

```{r}

figS13 <- figS13A+figS13B
figS13

#save output
#ggsave("figures/figS13.pdf",plot=figS13,device="pdf",useDingbats=FALSE,width=10)

```

### PERMANOVA

Altogether & across regions during the mid-Holocene

```{r}

#Run PERMANOVA
dis_mat_disagg <- vegdist(dd_pan_wide_disagg[,9:15], method="horn")

adonis2(dis_mat_disagg~dd_pan_wide_disagg$age_group_region,permutations=9999)

adonis2(dis_mat_disagg~dd_pan_wide_disagg$age_group+dd_pan_wide_disagg$region,permutations=9999)


#Mid-Holocene PERMANOVA
dd_pan_wide_holocene_disagg <- dd_pan_wide_disagg %>% 
  filter(age_group=="pre_exp")

dis_mat_holocene_disagg <- vegdist(dd_pan_wide_holocene_disagg[,9:15],method="horn")

adonis2(dis_mat_holocene_disagg~dd_pan_wide_holocene_disagg$region,permutations=9999)



#Testing dispersion assumptions
mod_age <- betadisper(dis_mat_disagg, dd_pan_wide_disagg$age_group_region)

boxplot(mod_age)
anova(mod_age)
permutest(mod_age)

```

Caribbean

```{r}

#Subset data and run PERMANOVA
dd_pan_wide_carib_disagg <- dd_pan_wide_disagg %>% 
  filter(region=="Bocas del Toro")

dis_mat_carib_disagg <- vegdist(dd_pan_wide_carib_disagg[,9:15],method="horn")

adonis2(dis_mat_carib_disagg~dd_pan_wide_carib_disagg$age_group + dd_pan_wide_carib_disagg$site_name,permutations=9999) 

adonis2(dis_mat_carib_disagg~dd_pan_wide_carib_disagg$age_group,permutations=9999)


#Testing dispersion assumptions
mod_age_carib <- betadisper(dis_mat_carib_disagg, dd_pan_wide_carib_disagg$age_group) 

mod_age_carib <- betadisper(dis_mat_carib_disagg, dd_pan_wide_carib_disagg$site_name) 

boxplot(mod_age_carib)
anova(mod_age_carib)
permutest(mod_age_carib)

```

Pacific

```{r}

#Subset data and run PERMANOVA 
dd_pan_wide_pacif_disagg <- dd_pan_wide_disagg %>% 
  filter(region=="Gulf of Panama")

dis_mat_pacif_disagg <- vegdist(dd_pan_wide_pacif_disagg[,9:15],method="horn")

adonis2(dis_mat_pacif_disagg~dd_pan_wide_pacif_disagg$age_group + dd_pan_wide_pacif_disagg$site_name,permutations=9999) 

adonis2(dis_mat_pacif_disagg~dd_pan_wide_pacif_disagg$age_group,permutations=9999,strata=dd_pan_wide_pacif_disagg$site_name)


#Testing dispersion assumptions
mod_age_pacif <- betadisper(dis_mat_pacif_disagg, dd_pan_wide_pacif_disagg$age_group) 

boxplot(mod_age_pacif)
anova(mod_age_pacif)
permutest(mod_age_pacif)

perm <- how(nperm = 9999)
setBlocks(perm) <- dd_pan_wide_pacif_disagg$site_name
permutest(mod_age_pacif, permutations = perm)

```

## Modeling proportional changes in pelagic and demersal denticles

Split by region and prepare for analysis.

```{r}

#Caribbean
dd_func_pan_carib <- dd_func_pan %>% 
  filter(basin=="Caribbean") %>%
  mutate(sample_age = factor(paste0(sample, " ", age_group)))%>% 
  droplevels()

#Pacific
dd_func_pan_pac <- dd_func_pan %>% 
  filter(basin=="Pacific") %>%
  mutate(sample_age = factor(paste0(sample, " ", age_group)))%>% 
  droplevels()

```

Build binomial models.

```{r}

#Caribbean pelagic
m1_carib_pelagic_binom <- glmmTMB(cbind(pelagic_count, demersal_count) ~ age_group + (1|sample_age),
              data=dd_func_pan_carib,
              family=binomial,
              REML=T)

#Caribbean demersal
m1_carib_demersal_binom <- glmmTMB(cbind(demersal_count, pelagic_count) ~ age_group + (1|sample_age),
              data=dd_func_pan_carib,
              family=binomial,
              REML=T)

#Pacific pelagic
m1_pac_pelagic_binom <- glmmTMB(cbind(pelagic_count, demersal_count) ~ age_group + (1|sample_age),
              data=dd_func_pan_pac,
              family=binomial,
              REML=T)

#Pacific demersal
m1_pac_demersal_binom <- glmmTMB(cbind(demersal_count, pelagic_count) ~ age_group + (1|sample_age),
              data=dd_func_pan_pac,
              family=binomial,
              REML=T)


#Assign to testing
model_to_test_carib <- m1_carib_pelagic_binom
model_to_test_pac <- m1_pac_pelagic_binom

```

Test model diagnostics.

```{r}

#Caribbean
model_sim_carib <- simulateResiduals(model_to_test_carib,n=1000)
plot(model_sim_carib)

#Pacific
model_sim_pac <- simulateResiduals(model_to_test_pac,n=1000)
plot(model_sim_pac)

```

Model summaries.

```{r}

summary(m1_carib_pelagic_binom)
summary(m1_carib_demersal_binom)
summary(m1_pac_pelagic_binom)
summary(m1_pac_demersal_binom)

```

Likelihood Ratio Tests.

```{r}

#Caribbean pelagic
m1_carib_pelagic_full <- glmmTMB(cbind(pelagic_count, demersal_count) ~ age_group + (1|sample_age),
              data=dd_func_pan_carib,
              family=binomial,
              REML=F)

m1_carib_pelagic_drop <- glmmTMB(cbind(pelagic_count, demersal_count) ~ (1|sample_age),
              data=dd_func_pan_carib,
              family=binomial,
              REML=F)

anova(m1_carib_pelagic_full,m1_carib_pelagic_drop) 


#Caribbean demersal
m1_carib_demersal_full <- glmmTMB(cbind(demersal_count, pelagic_count) ~ age_group + (1|sample_age),
              data=dd_func_pan_carib,
              family=binomial,
              REML=F)

m1_carib_demersal_drop <- glmmTMB(cbind(demersal_count, pelagic_count) ~ (1|sample_age),
              data=dd_func_pan_carib,
              family=binomial,
              REML=F)

anova(m1_carib_demersal_full,m1_carib_demersal_drop) 


#Pacific pelagic
m1_pac_pelagic_full <- glmmTMB(cbind(pelagic_count, demersal_count) ~ age_group + (1|sample_age),
              data=dd_func_pan_pac,
              family=binomial,
              REML=F)

m1_pac_pelagic_drop <- glmmTMB(cbind(pelagic_count, demersal_count) ~ (1|sample_age),
              data=dd_func_pan_pac,
              family=binomial,
              REML=F)

anova(m1_pac_pelagic_full,m1_pac_pelagic_drop) 


#Pacific demersal
m1_pac_demersal_full <- glmmTMB(cbind(demersal_count, pelagic_count) ~ age_group + (1|sample_age),
              data=dd_func_pan_pac,
              family=binomial,
              REML=F)

m1_pac_demersal_drop <- glmmTMB(cbind(demersal_count, pelagic_count) ~ (1|sample_age),
              data=dd_func_pan_pac,
              family=binomial,
              REML=F)

anova(m1_pac_demersal_full,m1_pac_demersal_drop)

```

Estimated marginal means.

```{r}

#Pacific
ef_pac_pelagic_binom <- effect("age_group", m1_pac_pelagic_binom)
ef_pac_demersal_binom <- effect("age_group", m1_pac_demersal_binom)

#Caribbean
ef_carib_pelagic_binom <- effect("age_group", m1_carib_pelagic_binom)
ef_carib_demersal_binom <- effect("age_group", m1_carib_demersal_binom)

```

## Modeling accumulation rates of pelagic and demersal denticles

Calculate counts and accumulation rates of pelagic and demersal denticle types.

```{r}

dd_func_pan_accum <- dd_func_pan_age %>% 
  mutate(pelagic_count = drag + abr_pel + rabr_pel,
         demersal_count = abr_dem + rabr_dem) %>% 
  mutate(pel_accum = pelagic_count/sed_weight_kg/no_years,
         dem_accum = demersal_count/sed_weight_kg/no_years)

```

Split by region and prepare for analysis.

```{r}

#Caribbean
dd_func_pan_accum_carib <- dd_func_pan_accum %>% 
  filter(basin=="Caribbean") %>%
  mutate(log_sed_weight_kg = log(sed_weight_kg)) %>% 
  mutate(site_age = factor(paste0(site_name, " ", age_group)),
         sample_age = factor(paste0(sample, " ", age_group))) %>% 
  droplevels()

#Pacific
dd_func_pan_accum_pac <- dd_func_pan_accum %>% 
  filter(basin=="Pacific") %>%
  mutate(log_sed_weight_kg = log(sed_weight_kg)) %>% 
  mutate(site_age = factor(paste0(site_name, " ", age_group)),
         sample_age = factor(paste0(sample, " ", age_group))) %>% 
  droplevels()

```

Using same model structure as for overall denticle counts.

```{r}

#Caribbean pelagic
m1_carib_pelagic <- glmmTMB(pelagic_count ~ age_group + log_sed_weight_kg + (1|sample_age)+ offset(log(no_years)),
              data=dd_func_pan_accum_carib,
              family=nbinom2,
              REML=T)

#Caribbean demersal
m1_carib_demersal <- glmmTMB(demersal_count ~ age_group + log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_func_pan_accum_carib,
              family=nbinom2,
              REML=T)

#Pacific pelagic
m1_pac_pelagic <- glmmTMB(pelagic_count ~ age_group + log_sed_weight_kg + (1|sample_age)+ offset(log(no_years)),
              data=dd_func_pan_accum_pac,
              family=nbinom2,
              REML=T)

#Pacific demersal
m1_pac_demersal <- glmmTMB(demersal_count ~ age_group + log_sed_weight_kg + (1|sample_age)+ offset(log(no_years)),
              data=dd_func_pan_accum_pac,
              family=nbinom2,
              REML=T)

#Assign to testing
model_to_test_carib <- m1_carib_pelagic 
model_to_test_pac <- m1_pac_pelagic

```

Source function to run conditional simulations.

```{r}
#| label: setup
#| include: false
source("functions/model_fit.R")

```

Test model diagnostics.

```{r}

#Caribbean
sim_data <- simulate_data(model_to_test_carib, nsim=5000)

dharma_resid <- createDHARMa(simulatedResponse=sim_data$sim,
                             observedResponse=dd_func_pan_accum_carib$pelagic_count,
                             fittedPredictedResponse=sim_data$mu)
plot(dharma_resid, quantreg=TRUE)


#Pacific
dd_func_pan_accum_pac_na <- dd_func_pan_accum_pac %>% 
  drop_na()

sim_data <- simulate_data(model_to_test_pac, nsim=5000)

dharma_resid <- createDHARMa(simulatedResponse=sim_data$sim,
                             observedResponse=dd_func_pan_accum_pac_na$pelagic_count,
                             fittedPredictedResponse=sim_data$mu)
plot(dharma_resid, quantreg=TRUE)

```

Model summaries.

```{r}

summary(m1_carib_pelagic)
summary(m1_carib_demersal)
summary(m1_pac_pelagic)
summary(m1_pac_demersal)

```

Likelihood Ratio Tests.

```{r}

#Caribbean pelagic
m1_carib_pelagic_full <- glmmTMB(pelagic_count ~ age_group + log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_func_pan_accum_carib,
              family=nbinom2,
              REML=F)

m1_carib_pelagic_drop <- glmmTMB(pelagic_count ~ log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_func_pan_accum_carib,
              family=nbinom2,
              REML=F)

anova(m1_carib_pelagic_full,m1_carib_pelagic_drop) 


#Caribbean demersal
m1_carib_demersal_full <- glmmTMB(demersal_count ~ age_group + log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_func_pan_accum_carib,
              family=nbinom2,
              REML=F)

m1_carib_demersal_drop <- glmmTMB(demersal_count ~ log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_func_pan_accum_carib,
              family=nbinom2,
              REML=F)

anova(m1_carib_demersal_full,m1_carib_demersal_drop) 


#Pacific pelagic
m1_pac_pelagic_full <- glmmTMB(pelagic_count ~ age_group + log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_func_pan_accum_pac,
              family=nbinom2,
              REML=F)

m1_pac_pelagic_drop <- glmmTMB(pelagic_count ~ log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_func_pan_accum_pac,
              family=nbinom2,
              REML=F)

anova(m1_pac_pelagic_full,m1_pac_pelagic_drop) 


#Pacific demersal
m1_pac_demersal_full <- glmmTMB(demersal_count ~ age_group + log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_func_pan_accum_pac,
              family=nbinom2,
              REML=F)

m1_pac_demersal_drop <- glmmTMB(demersal_count ~ log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_func_pan_accum_pac,
              family=nbinom2,
              REML=F)

anova(m1_pac_demersal_full,m1_pac_demersal_drop) 

```

Estimated marginal means.

```{r}

#predict for 100 years and 1kg sediment as before
yr_predict <-100 
offset_value_predict <- log(yr_predict)


#Pacific

ef_pac_pelagic <- effect("age_group", m1_pac_pelagic,
                         given.values=c(log_sed_weight_kg=0), #1kg
                         offset=offset_value_predict) 

ef_pac_demersal <- effect("age_group", m1_pac_demersal,
                         given.values=c(log_sed_weight_kg=0), #1kg
                         offset=offset_value_predict) 

#Caribbean

ef_carib_pelagic <- effect("age_group", m1_carib_pelagic,
                         given.values=c(log_sed_weight_kg=0), #1kg
                         offset=offset_value_predict) 

ef_carib_demersal <- effect("age_group", m1_carib_demersal,
                         given.values=c(log_sed_weight_kg=0), #1kg
                         offset=offset_value_predict) 


#Combine for plotting
ef_pac_df_pelagic <- as.data.frame(ef_pac_pelagic) %>% 
  add_column(basin='Pacific',.before="age_group") %>%
  add_column(region='Gulf of Panama',.before="age_group") %>% 
  add_column(life_mode='Pelagic',.before="age_group")

ef_pac_df_demersal <- as.data.frame(ef_pac_demersal) %>% 
  add_column(basin='Pacific',.before="age_group") %>%
  add_column(region='Gulf of Panama',.before="age_group") %>% 
  add_column(life_mode='Demersal',.before="age_group")

ef_carib_df_pelagic <- as.data.frame(ef_carib_pelagic)%>% 
  add_column(basin='Caribbean',.before="age_group") %>%
  add_column(region='Bocas del Toro',.before="age_group")%>% 
  add_column(life_mode='Pelagic',.before="age_group")

ef_carib_df_demersal <- as.data.frame(ef_carib_demersal)%>% 
  add_column(basin='Caribbean',.before="age_group") %>%
  add_column(region='Bocas del Toro',.before="age_group")%>% 
  add_column(life_mode='Demersal',.before="age_group")

ef_pan_df_lifemode <- rbind(ef_pac_df_pelagic,ef_pac_df_demersal,ef_carib_df_pelagic,ef_carib_df_demersal) %>% 
  mutate(age_group = fct_relevel(age_group,c("pre_exp","modern")))

```

### figure S14

```{r}

figS14 <- ggplot(ef_pan_df_lifemode, aes(x=life_mode, y=fit,shape=age_group,fill=basin)) +
  geom_pointrange(aes(ymin=fit-se, ymax=fit+se,shape=age_group,fill=basin),color='black',fatten=4,size=1.5, position=position_jitterdodge(seed=10)) + 
  theme_bw()+
  labs(y="Denticle accumulation per kg sediment per century",x="Ecological group",shape="Time period",fill="Region") +
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Recent"))+
  scale_x_discrete(limits=c("Pelagic","Demersal"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  theme(panel.grid.minor = element_blank())+
  theme(legend.position = "right")+
  guides(fill="none")+
  theme(aspect.ratio=1.5)+
  facet_wrap(~basin,scales="free")+
  theme(strip.background = element_rect(fill = NA, color = NA),
        strip.text = element_text(size = 16))
figS14


```

Save output (fig S13)

```{r}

#ggsave("figures/figS14.png",plot=figS14,device="png",width=10)

```

Magnitude of change, based on the models.

```{r}

#Caribbean
modern_carib_pel <- ef_pan_df_lifemode[6,5]
holocene_carib_pel <- ef_pan_df_lifemode[5,5]

modern_carib_dem <- ef_pan_df_lifemode[8,5]
holocene_carib_dem <- ef_pan_df_lifemode[7,5]

(modern_carib_pel-holocene_carib_pel)/holocene_carib_pel*100 
(modern_carib_dem-holocene_carib_dem)/holocene_carib_dem*100 


#Pacific
modern_pac_pel <- ef_pan_df_lifemode[2,5]
holocene_pac_pel <- ef_pan_df_lifemode[1,5]

modern_pac_dem <- ef_pan_df_lifemode[4,5]
holocene_pac_dem <- ef_pan_df_lifemode[3,5]

(modern_pac_pel-holocene_pac_pel)/holocene_pac_pel*100 
(modern_pac_dem-holocene_pac_dem)/holocene_pac_dem*100 
```

## Denticle taphonomy

### Read in raw data

```{r}

taph_raw <- read.csv("data/dd_taph.csv",header=T,na.strings=".")

```

### Clean and merge with denticle count data

```{r}

#Change taphonomic grades to numerical scores
taph <- taph_raw %>% 
  mutate(frag_base_score=case_when(
    frag_base=="A" ~ 0,
    frag_base=="B" ~1,
    frag_base=="C"~2)) %>% 
  mutate(frag_crown_score=case_when(
    frag_crown=="A" ~ 0,
    frag_crown=="B" ~1,
    frag_crown=="C"~2,
    frag_crown=="D"~3)) %>% 
  mutate(frag_peaks_score=case_when(
    frag_peaks=="A" ~ 0,
    frag_peaks=="B" ~ 1)) %>% 
  mutate(surf_alt_score=case_when(
    surf_alt=="A" ~ 0,
    surf_alt=="B" ~ 1,
    surf_alt=="C" ~ 2,
    surf_alt=="D" ~ 3)) %>%
  mutate(discolor_score=case_when(
    discolor=="A" ~ 0,
    discolor=="B" ~ 1,
    discolor=="C" ~ 2)) %>%
  mutate(tot_frag_score=(frag_base_score/2)+(frag_crown_score/3)+frag_peaks_score,
         tot_score=tot_frag_score+(surf_alt_score/3)+(discolor_score/2))%>% 
  filter(func_morph != "RAY") #remove any ray denticles


#Merge with denticle counts and sample metadata
taph_age <- taph %>% 
left_join(dd_pan_age_clean_pre %>% select(unique,total_dd_count,dd_kg,no_years,age_group),by="unique") %>%
  mutate(age_group = if_else(unique == "JC22-2-2-135", "pre_exp", age_group)) %>% 
  mutate(age_group = fct_relevel(age_group,c("pre_exp","modern")))%>%
  filter(age_group %in% c("modern","pre_exp"))


#Calculate means for each sample
taph_summ <- taph_age %>% 
  group_by(basin, region, site_name, sample, unique,dd_kg, age_group,no_years) %>% 
  summarize(mean_weathering = mean(tot_score,na.rm=T),
            mean_frag = mean(tot_frag_score,na.rm=T),
            frag_count = length(which(is_frag == "Y")),
            frag_prop = length(which(is_frag == "Y"))/length(unique))

```

### figure S8

Across regions (fig. S8A).

```{r}

figS8A <- ggplot(taph_summ,aes(x=region, y=mean_weathering, fill=region, shape=age_group))+
  geom_point(aes(shape=age_group),color="gray70",fill="gray70",alpha=0.4,size=2,position=position_jitterdodge(jitter.width=0.5,dodge.width=0.7,seed=10))+
  geom_boxplot(aes(fill=region),alpha=0.8,outlier.shape = NA,width=0.5,position=position_dodge(width=0.7))+
  theme_bw(base_size=12)+
  labs(y="Weathering score",x="Region",fill=" ",shape="Age group") +
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Sub Recent"))+
  scale_x_discrete(limits=c("Bocas del Toro","Gulf of Panama"),labels=c("Caribbean\n(Bocas del Toro)","Pacific\n(Gulf of Panama)"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  ylim(0,3)+
  theme(legend.position = "none")+
  theme(aspect.ratio=1.5)
figS8A

```

Across sites (fig. S8B).

```{r}

#Regional means
regional_weathering_means <- taph_summ %>%
  group_by(region) %>% 
  summarize(region_mean = mean(mean_weathering,na.rm=T))
regional_weathering_means

#Site means and SD
site_taph_means <- taph_summ %>% 
  mutate(site_name2 = case_when(
    site_name == "Sweet Bocas5" ~ "Sweet Bocas",
    site_name == "Sweet Bocas6" ~ "Sweet Bocas",
    site_name == "Sweet Bocas7" ~ "Sweet Bocas",
    site_name == "Sweet Bocas9" ~ "Sweet Bocas",
    site_name == "Sweet Bocas10" ~ "Sweet Bocas",
    TRUE~site_name
  )) %>% 
  group_by(region,site_name2,age_group) %>% 
  summarize(mean_score = mean(mean_weathering,na.rm=T),
            sd_score = sd(mean_weathering,na.rm=T),
            se_score = sd(mean_weathering,na.rm=T)/sqrt(length(mean_weathering))) %>% 
  rename(site_name=site_name2)
site_taph_means

#Plot
figS8B <- ggplot(site_taph_means,aes(x=site_name, y=mean_score, shape=age_group))+
  geom_hline(yintercept=regional_weathering_means$region_mean[2],color="deepskyblue3",linetype=2,linewidth=0.8)+
  geom_hline(yintercept=regional_weathering_means$region_mean[1],color="darkgoldenrod2",linetype=2,linewidth=0.8)+
  geom_pointrange(aes(ymin=mean_score-sd_score, ymax=mean_score+sd_score,shape=age_group,fill=region),color='black',fatten=3.5,size=1.5, position=position_jitterdodge(seed=10)) +
  theme_bw()+
  labs(y="Weathering score",x="Site",fill="Region",shape="Time period") +
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Recent"))+
  scale_x_discrete(limits=c("Sweet Bocas","Casa Blanca","Cayo Adriana","Punta Donato","Isla Iguana","Saboga","Contadora"),labels=c("Sweet\nBocas","Casa\nBlanca","Cayo\nAdriana","Punta\nDonato","Isla\nIguana","Saboga","Contadora"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  guides(fill="none")+
  theme(axis.title.y=element_blank())+
  ylim(0,3)+
  theme(legend.position = "bottom")+
  theme(aspect.ratio=0.9)
figS8B

```

By denticle morphotype (fig. S8C).

```{r}

#Period means and SD
age_taph_means <- taph_age %>% 
  group_by(age_group,func_morph) %>% 
  summarize(mean_score = mean(tot_score,na.rm=T),
            sd_score = sd(tot_score,na.rm=T),
            se_score = sd(tot_score,na.rm=T)/sqrt(length(tot_score))) 
age_taph_means

#Plot
figS8C <- age_taph_means %>% 
  filter(func_morph %in% c("A","RA","DR","ABR","DEFENSE","GF")) %>%
ggplot(aes(x=func_morph, y=mean_score,shape=age_group))+
geom_pointrange(aes(ymin=mean_score-sd_score, ymax=mean_score+sd_score,shape=age_group),fill="gray50",color='black',fatten=3.5,size=1.5, position=position_jitterdodge(seed=10)) +
  theme_bw()+
  labs(y="Weathering score",x="Functional morphotype",shape="Time period") +
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Recent"))+
  scale_x_discrete(limits=c("DR","RA","A","GF","DEFENSE"),labels=c("Drag","Abrasion","Ridged\nAbrasion","Gen\nFunc","Defense"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  ylim(0,3)+
  theme(axis.title.y=element_blank())+
  theme(legend.position = "none")+
  theme(aspect.ratio=1.1)
figS8C


```

Combine and save output (fig. S8).

```{r}

figS8 <- (figS8A+figS8B+figS8C)+
  plot_annotation(tag_levels = 'A')&
  theme(
    plot.tag = element_text(size = 18, face = "bold"))
figS8

#save output
#ggsave("figures/figS8.pdf",plot=figS8,device="pdf",useDingbats=FALSE,width=14)

```

### figure S9

Denticle preservation vs. abundance

```{r}

figS9A <- ggplot(taph_summ, aes(y=mean_weathering,x=dd_kg))+
  geom_point(aes(fill=basin, shape=age_group),color='gray30',size=2.5,alpha=0.8)+
  labs(y="Weathering score",x="Denticles per kg sediment",shape="Time period",fill="Region")+
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Recent"))+
  theme_bw()+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans"))+
  ylim(0,3)+
  guides(fill="none")+
  scale_x_continuous(trans='log10')+
  annotation_logticks(sides="b")+
  theme(legend.position = "none")+
  theme(aspect.ratio = 1)+
  facet_wrap(~basin)+
  theme(strip.background = element_rect(fill = NA, color = NA),
        strip.text = element_text(size = 16))
figS9A

```

Denticle preservation vs. time-averaging

```{r}

figS9B <- taph_summ %>%
  ggplot(aes(y=mean_weathering,x=no_years))+
  geom_point(aes(fill=basin, shape=age_group),color='gray30',size=2.5,alpha=0.8)+
  labs(y="Weathering score",x="Number of years per sample",shape="Time period",fill="Region")+
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Recent"))+
  theme_bw()+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans"))+
  ylim(0,3)+
  guides(fill="none")+
  scale_x_continuous(trans='log10')+
  annotation_logticks(sides="b")+
  theme(legend.position = "bottom")+
  theme(aspect.ratio = 1)+
  facet_wrap(~basin)+
  theme(strip.background = element_rect(fill = NA, color = NA),
        strip.text = element_text(size = 16))
figS9B

```

Combine and save output (fig. S9).

```{r warning=F}

figS9 <- figS9A/figS9B+
  plot_layout(widths = c(3,-8,3))+
  plot_annotation(tag_levels = 'A')&
  theme(
    plot.tag = element_text(size = 18, face = "bold"))
  figS9

#save output
#ggsave("figures/figS9.png",plot=figS9,device="png",width=8,height=9)

```
