---
title: "Cross-isthmian denticle abundances"
format: html
editor: visual
---

## Overview

Here, we model denticle accumulation rates to make comparisons between time periods and regions. This script cleans and filters the raw denticle count data, runs the analysis, and plots Figure 1 and the related supplementary figures.

Author: Erin Dillon, emdillon23\@gmail.com

## Libraries

```{r}
#| warning: false

library(glmmTMB)
library(DHARMa)
library(effects)
library(emmeans)
library(psych)
library(tidyverse)
library(ggridges)
library(patchwork)

```

## Data

### Read in raw data

```{r}

#BChron age models for each core
bchron_age_depth <- read.csv("data/bchron_ages.csv",header=T)

#Sample intervals
sample_interval <- read.csv("data/sampling_intervals_pacific.csv")

#Denticle counts
dd_pan <- read.csv("data/cross_isthmus_dd.csv",header=T,na.strings=".")

```

### Data cleaning

Map sample midpoint predictions from BChron onto sampling interval labels

```{r}

bchron_age_depth <- bchron_age_depth %>% 
  #join with sample labels
  left_join(sample_interval,by=c("region","date_locality","unique","sample","depth")) %>% 
  #rename a couple of variables for clarity
  rename(midpoint=depth) %>% 
  rename(depth=depth_lower) %>% 
  rename(date_unique=unique) %>% 
  rename(unique=sampling_unique) %>% 
  rename(median_num_years_per_sample=perc_50)

```

Match up denticle counts with age-depth model estimates

```{r}

dd_pan_age <- dd_pan %>% 
  #combining denticle counts with age estimates
  left_join(bchron_age_depth,by=c("region","sample","depth","unique")) %>% 
  mutate(basin = as.factor(basin),
         region = as.factor(region),
         site_name = as.factor(site_name),
         sample = as.factor(sample)) %>%
  #estimated amount of time in each sample
  mutate(no_years=case_when(
  #bulk sample age estimates
    basin=="Caribbean" ~ num_yrs_est_bulks, 
  #median from age-depth models
    basin=="Pacific" ~ median_num_years_per_sample)) %>%
  #25th percentile, for plotting
  mutate(no_years_25=case_when(
    basin=="Caribbean" ~ num_yrs_est_bulks_lower, 
    basin=="Pacific" ~ perc_25)) %>% 
  #75th percentile, for plotting
  mutate(no_years_75=case_when(
    basin=="Caribbean" ~ num_yrs_est_bulks_upper, 
    basin=="Pacific" ~ perc_75)) %>% 
  #calculate raw denticle accumulation metrics
  mutate(dd_kg=total_dd_count/sed_weight_kg,
         dd_kg_yr=dd_kg/no_years,
         dd_kg_yr_25=dd_kg/no_years_25, 
         dd_kg_yr_75=dd_kg/no_years_75) %>% 
  #using median age for each sampling interval
  rename(age_AD_BC=predicted_age_median, 
         age_BP=predicted_age_BP_median) %>% 
  #filtering out one site with difficult to interpret time-averaging
  filter(site_name != "Airport Point")

```

Filtering to just core tops/bottoms to facilitate comparability across the isthmus

```{r}

dd_pan_age_clean_pre <- dd_pan_age %>%
  #applying age cut-offs
  mutate(age_group = case_when(
      age_AD_BC > 1900 ~ "modern", #last ~100 years
      #Holocene, site-specific lower cut-offs given differences in accretion histories, and aiming for similar temporal range across basins
      site_name == "Contadora" & age_AD_BC < -2000 ~ "pre_exp",  
      site_name == "Isla Iguana" & age_AD_BC < -2000 ~ "pre_exp",
      site_name == "Saboga" & age_AD_BC < -1000 ~ "pre_exp", 
      basin == "Caribbean" & type == "modern_bulk" ~ "modern", 
      basin == "Caribbean" & type == "Holocene_bulk" ~ "pre_exp")) %>%
  mutate(age_group = fct_relevel(age_group,c("pre_exp","modern")))%>%
  #keeping just the relevant samples for each time period
  filter(age_group %in% c("modern","pre_exp")) %>%
  droplevels() %>% 
  #adding a column for the sampling year
  mutate(sampling_year = case_when(site_name == "Contadora" ~ 2018,
                                   site_name == "Isla Iguana" ~ 2022,
                                   site_name == "Saboga" ~ 2019,
                                   region == "Bocas del Toro" ~ 2014))

```

Save data prior to doing any cleaning related with reef matrix accumulation.

```{r}

#write.csv(dd_pan_age_clean_pre,"clean/dd_pan_age_clean_pre.csv")

```

Aggregating core tops into composite "bulk" samples given the presence of looser framework and less compaction. Aim here is to make the top \~15cm usable vs. filtering it out.

```{r}

dd_pan_age_clean_BULK <- dd_pan_age_clean_pre %>% 
  #filter to the top intervals 0-15cm
  filter(basin=="Pacific" & depth<20) %>%
  group_by(basin, region, site_name,sample) %>% 
  #summarize sample info for these top intervals
  summarize(dd_top = sum(total_dd_count,na.rm=T),
            weight_top = sum(sed_weight_kg,na.rm=T),
            years_top = sampling_year-min(age_AD_BC)) %>% #median date lower sample
  unique() %>% 
  mutate(dd_kg_yr_top = dd_top/weight_top/years_top)

```

Now replace the top sampling intervals with each of these analytically pooled "bulk" samples.

```{r}

dd_pan_age_clean <- dd_pan_age_clean_pre %>%
  #remove just the rows pooled into the "bulk" samples
  filter(depth>12 | is.na(depth)==T) %>% 
  #impute pooled sample info at depth=15, from  `dd_pan_age_clean_BULK` while retaining other columns for depth=15
  mutate(total_dd_count = case_when( #aggregated dd count
    depth==15 & sample=="BF18-1-11" ~ dd_pan_age_clean_BULK$dd_top[dd_pan_age_clean_BULK$sample=="BF18-1-11"],
    depth==15 & sample=="BF18-1-12" ~ dd_pan_age_clean_BULK$dd_top[dd_pan_age_clean_BULK$sample=="BF18-1-12"],
    depth==15 & sample=="BF18-1-13" ~ dd_pan_age_clean_BULK$dd_top[dd_pan_age_clean_BULK$sample=="BF18-1-13"],
    depth==15 & sample=="AT19-1-1" ~ dd_pan_age_clean_BULK$dd_top[dd_pan_age_clean_BULK$sample=="AT19-1-1"],
    depth==15 & sample=="AT19-1-2" ~ dd_pan_age_clean_BULK$dd_top[dd_pan_age_clean_BULK$sample=="AT19-1-2"],
    depth==15 & sample=="AT19-1-3" ~ dd_pan_age_clean_BULK$dd_top[dd_pan_age_clean_BULK$sample=="AT19-1-3"],
    depth==15 & sample=="JC22-2-1" ~ dd_pan_age_clean_BULK$dd_top[dd_pan_age_clean_BULK$sample=="JC22-2-1"],
    depth==15 & sample=="JC22-2-2" ~ dd_pan_age_clean_BULK$dd_top[dd_pan_age_clean_BULK$sample=="JC22-2-2"],
    depth==15 & sample=="JC22-2-3" ~ dd_pan_age_clean_BULK$dd_top[dd_pan_age_clean_BULK$sample=="JC22-2-3"],
    TRUE ~ total_dd_count)) %>% 
  mutate(sed_weight_kg = case_when( #aggregated sediment weight
    depth==15 & sample=="BF18-1-11" ~ dd_pan_age_clean_BULK$weight_top[dd_pan_age_clean_BULK$sample=="BF18-1-11"],
    depth==15 & sample=="BF18-1-12" ~ dd_pan_age_clean_BULK$weight_top[dd_pan_age_clean_BULK$sample=="BF18-1-12"],
    depth==15 & sample=="BF18-1-13" ~ dd_pan_age_clean_BULK$weight_top[dd_pan_age_clean_BULK$sample=="BF18-1-13"],
    depth==15 & sample=="AT19-1-1" ~ dd_pan_age_clean_BULK$weight_top[dd_pan_age_clean_BULK$sample=="AT19-1-1"],
    depth==15 & sample=="AT19-1-2" ~ dd_pan_age_clean_BULK$weight_top[dd_pan_age_clean_BULK$sample=="AT19-1-2"],
    depth==15 & sample=="AT19-1-3" ~ dd_pan_age_clean_BULK$weight_top[dd_pan_age_clean_BULK$sample=="AT19-1-3"],
    depth==15 & sample=="JC22-2-1" ~ dd_pan_age_clean_BULK$weight_top[dd_pan_age_clean_BULK$sample=="JC22-2-1"],
    depth==15 & sample=="JC22-2-2" ~ dd_pan_age_clean_BULK$weight_top[dd_pan_age_clean_BULK$sample=="JC22-2-2"],
    depth==15 & sample=="JC22-2-3" ~ dd_pan_age_clean_BULK$weight_top[dd_pan_age_clean_BULK$sample=="JC22-2-3"],
    TRUE ~ sed_weight_kg)) %>%
  mutate(no_years = case_when( #estimated number of years
    depth==15 & sample=="BF18-1-11" ~ dd_pan_age_clean_BULK$years_top[dd_pan_age_clean_BULK$sample=="BF18-1-11"],
    depth==15 & sample=="BF18-1-12" ~ dd_pan_age_clean_BULK$years_top[dd_pan_age_clean_BULK$sample=="BF18-1-12"],
    depth==15 & sample=="BF18-1-13" ~ dd_pan_age_clean_BULK$years_top[dd_pan_age_clean_BULK$sample=="BF18-1-13"],
    depth==15 & sample=="AT19-1-1" ~ dd_pan_age_clean_BULK$years_top[dd_pan_age_clean_BULK$sample=="AT19-1-1"],
    depth==15 & sample=="AT19-1-2" ~ dd_pan_age_clean_BULK$years_top[dd_pan_age_clean_BULK$sample=="AT19-1-2"],
    depth==15 & sample=="AT19-1-3" ~ dd_pan_age_clean_BULK$years_top[dd_pan_age_clean_BULK$sample=="AT19-1-3"],
    depth==15 & sample=="JC22-2-1" ~ dd_pan_age_clean_BULK$years_top[dd_pan_age_clean_BULK$sample=="JC22-2-1"],
    depth==15 & sample=="JC22-2-2" ~ dd_pan_age_clean_BULK$years_top[dd_pan_age_clean_BULK$sample=="JC22-2-2"],
    depth==15 & sample=="JC22-2-3" ~ dd_pan_age_clean_BULK$years_top[dd_pan_age_clean_BULK$sample=="JC22-2-3"],
    TRUE ~ no_years)) %>% 
  mutate(dd_kg_yr = case_when( #dd accumulation in pooled sample
    depth==15 & sample=="BF18-1-11" ~ dd_pan_age_clean_BULK$dd_kg_yr_top[dd_pan_age_clean_BULK$sample=="BF18-1-11"],
    depth==15 & sample=="BF18-1-12" ~ dd_pan_age_clean_BULK$dd_kg_yr_top[dd_pan_age_clean_BULK$sample=="BF18-1-12"],
    depth==15 & sample=="BF18-1-13" ~ dd_pan_age_clean_BULK$dd_kg_yr_top[dd_pan_age_clean_BULK$sample=="BF18-1-13"],
    depth==15 & sample=="AT19-1-1" ~ dd_pan_age_clean_BULK$dd_kg_yr_top[dd_pan_age_clean_BULK$sample=="AT19-1-1"],
    depth==15 & sample=="AT19-1-2" ~ dd_pan_age_clean_BULK$dd_kg_yr_top[dd_pan_age_clean_BULK$sample=="AT19-1-2"],
    depth==15 & sample=="AT19-1-3" ~ dd_pan_age_clean_BULK$dd_kg_yr_top[dd_pan_age_clean_BULK$sample=="AT19-1-3"],
    depth==15 & sample=="JC22-2-1" ~ dd_pan_age_clean_BULK$dd_kg_yr_top[dd_pan_age_clean_BULK$sample=="JC22-2-1"],
    depth==15 & sample=="JC22-2-2" ~ dd_pan_age_clean_BULK$dd_kg_yr_top[dd_pan_age_clean_BULK$sample=="JC22-2-2"],
    depth==15 & sample=="JC22-2-3" ~ dd_pan_age_clean_BULK$dd_kg_yr_top[dd_pan_age_clean_BULK$sample=="JC22-2-3"],
    TRUE ~ dd_kg_yr))

```

Filtering out samples with anomalously high reef accretion rates compared with previous work in the region. Loose framework at top of cores potentially creating inflated time estimates (see Toth et al. 2012).

```{r}

dd_pan_age_clean <- dd_pan_age_clean %>%
  filter(no_years>5) #cutoff = 1cm/yr (10m/1000y), vs. previous work showing accretion rates within 5m/1000y (e.g., Toth et al. 2012)

```

Adding a couple of columns for the models we'll build below

```{r}

dd_pan_age_clean <- dd_pan_age_clean %>%
  mutate(log_sed_weight_kg = log(sed_weight_kg),
         log_no_years = log(no_years),
         site_age = factor(paste0(site_name, " ", age_group)),
         sample_age = factor(paste0(sample, " ", age_group)))

```

Save the clean data

```{r}

#write.csv(dd_pan_age_clean,"clean/dd_pan_age_clean.csv")

```

### Sample sizes for analysis

```{r}

#samples by basin, after cleaning and filtering
table(dd_pan_age_clean$basin)

#total denticle count
sum(dd_pan_age_clean$total_dd_count,na.rm=T)

#mean no_years per sample
mean(dd_pan_age_clean$no_years,na.rm=T)

#denticle count per basin
dd_pan_age_clean %>% 
  group_by(basin) %>% 
  summarize(dd = sum(total_dd_count,na.rm=T))

#denticle count by basin and time period
dd_pan_age_clean %>% 
  group_by(basin,age_group) %>% 
  summarize(dd = sum(total_dd_count,na.rm=T),
            n = n())

#sample sizes by basin, site, and time period for table S2
dd_pan_age_clean %>% 
  group_by(basin,age_group,site_name) %>% 
  summarize(dd = sum(total_dd_count,na.rm=T),
            kg = sum(sed_weight_kg,na.rm=T),
            n = n())

```

## Modeling denticle accumulation rates

Split the data by basin, given the differences in sampling and data structures.

```{r}

dd_pan_carib <- dd_pan_age_clean %>% 
  filter(basin=="Caribbean") %>% 
  droplevels()

dd_pan_pac <- dd_pan_age_clean %>% 
  filter(basin=="Pacific")%>% 
  droplevels() %>% 
  drop_na(total_dd_count)

```

View relevant variables.

```{r}

#Caribbean
pairs.panels(dd_pan_carib %>% select(total_dd_count,log_sed_weight_kg,log_no_years,age_group))

#Pacific
pairs.panels(dd_pan_pac %>% select(total_dd_count,log_sed_weight_kg,log_no_years,age_group))

```

### Caribbean

Fit models.

```{r}

#testing several model structures

m1_carib <- glmmTMB(total_dd_count ~ age_group + log_sed_weight_kg + (1|site_name/unique) + offset(log(no_years)), 
                    data=dd_pan_carib,
                    family=poisson,
                    REML=T)


m2_carib <- glmmTMB(total_dd_count ~ age_group + log_sed_weight_kg + (1|site_name) + offset(log(no_years)), 
                    data=dd_pan_carib,
                    family=nbinom2,
                    REML=T)


m3_carib <- glmmTMB(total_dd_count ~ age_group + log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_pan_carib,
              family=nbinom2,
              REML=T)


m4_carib <- glmmTMB(total_dd_count ~ age_group + log_sed_weight_kg + (1|sample_age/unique) + offset(log(no_years)),
              data=dd_pan_carib,
              family=nbinom2,
              REML=T)


#assign to testing
model_to_test_carib <- m3_carib

```

Test model diagnostics.

```{r}

#DHARMa

model_sim_carib <- simulateResiduals(model_to_test_carib,n=1000)
plot(model_sim_carib)
testDispersion(model_sim_carib)

testCategorical(model_sim_carib,catPred=dd_pan_carib$age_group)
testQuantiles(model_sim_carib,predictor=dd_pan_carib$log_sed_weight_kg)
testQuantiles(model_sim_carib,predictor=log(dd_pan_carib$no_years))

```

Source function to run conditional simulations.

```{r}
#| label: setup
#| include: false
source("functions/model_fit.R")

```

```{r}

#Conditional simulations

sim_data <- simulate_data(model_to_test_carib, nsim=5000)

dharma_resid <- createDHARMa(simulatedResponse=sim_data$sim,
                             observedResponse=dd_pan_carib$total_dd_count,
                             fittedPredictedResponse=sim_data$mu)

plot(dharma_resid, quantreg=TRUE)
plotResiduals(dharma_resid, form=dd_pan_carib$log_sed_weight_kg)
plotResiduals(dharma_resid, form=dd_pan_carib$site_age)
plotResiduals(dharma_resid, form=dd_pan_carib$age_group)
plotResiduals(dharma_resid, form=log(dd_pan_carib$no_years))

```

Compare models.

```{r}

AIC(m1_carib,m2_carib,m3_carib,m4_carib) #M3 has lowest AIC

```

Model summary.

```{r}

model_selected_carib <- m3_carib

summary(model_selected_carib)

```

Likelihood Ratio Test to obtain p-value and test statistic.

```{r}

#full model
m3_carib_full <- glmmTMB(total_dd_count ~ age_group + log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_pan_carib,
              family=nbinom2,
              REML=F)


#same model structure, with age_group dropped
m3_carib_drop <- glmmTMB(total_dd_count ~ log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_pan_carib,
              family=nbinom2,
              REML=F)


#compare, statistically significant difference
anova(m3_carib_full,m3_carib_drop)

```

### Pacific

Fit models.

```{r}

#testing several model structures

m1_pac <- glmmTMB(total_dd_count ~ age_group + log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_pan_pac,
              family=nbinom2,
              REML=T)


m2_pac <- glmmTMB(total_dd_count ~ age_group + log_sed_weight_kg + (1|site_age) + offset(log(no_years)),
              data=dd_pan_pac,
              family=nbinom2,
              REML=T)


m3_pac <- glmmTMB(total_dd_count ~ age_group + log_sed_weight_kg + (1|sample_age/unique) + offset(log(no_years)),
              data=dd_pan_pac,
              family=nbinom2,
              REML=T) 


m4_pac <- glmmTMB(total_dd_count ~ age_group + log_sed_weight_kg + (1|site_name/sample/unique) + offset(log(no_years)),
                  data=dd_pan_pac,
                  family=poisson,
                  REML=T)


m5_pac <- glmmTMB(total_dd_count ~ age_group + log_sed_weight_kg + (1|site_name/sample) + offset(log(no_years)),
                  data=dd_pan_pac,
                  family=poisson,
                  REML=T)


m6_pac <- glmmTMB(total_dd_count ~ age_group + log_sed_weight_kg + (1|site_name/sample) + offset(log(no_years)),
                  data=dd_pan_pac,
                  family=nbinom2,
                  REML=T)


m7_pac <- glmmTMB(total_dd_count ~ age_group + log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_pan_pac,
              disp=~age_group,
              family=nbinom2,
              REML=T)


#assign to testing
model_to_test_pac <- m1_pac

```

Test model diagnostics.

```{r}

#DHARMa

model_sim_pac <- simulateResiduals(model_to_test_pac,n=1000)
plot(model_sim_pac)
testDispersion(model_sim_pac)

testCategorical(model_sim_pac,catPred=dd_pan_pac$age_group)
testQuantiles(model_sim_pac,predictor=dd_pan_pac$log_sed_weight_kg)
testQuantiles(model_sim_pac,predictor=log(dd_pan_pac$no_years))


```

```{r}

#Conditional simulations

sim_data <- simulate_data(model_to_test_pac, nsim=5000)

dharma_resid <- createDHARMa(simulatedResponse=sim_data$sim,
                             observedResponse=dd_pan_pac$total_dd_count,
                             fittedPredictedResponse=sim_data$mu)

plot(dharma_resid, quantreg=TRUE)
plotResiduals(dharma_resid, form=dd_pan_pac$log_sed_weight_kg)
plotResiduals(dharma_resid, form=dd_pan_pac$site_age)
plotResiduals(dharma_resid, form=dd_pan_pac$age_group)
plotResiduals(dharma_resid, form=log(dd_pan_pac$no_years))

```

Test temporal autocorrelation.

```{r}

fit_sim_time <- recalculateResiduals(model_sim_pac, group = dd_pan_pac$age_AD_BC)
testTemporalAutocorrelation(fit_sim_time, time = unique(dd_pan_pac$age_AD_BC))

#no significant temporal autocorrelation

```

Compare models.

```{r}

AIC(m1_pac,m2_pac,m3_pac,m4_pac,m5_pac,m6_pac,m7_pac) #M1 has lowest AIC (followed by M3 and M7)

```

Model summary.

```{r}

model_selected_pac <- m1_pac

summary(model_selected_pac)

```

Likelihood Ratio Test to obtain p-value and test statistic.

```{r}

#full model
m1_pac_full <- glmmTMB(total_dd_count ~ age_group + log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_pan_pac,
              family=nbinom2,
              REML=F)


#same model structure, with age_group dropped
m1_pac_drop <- glmmTMB(total_dd_count ~ log_sed_weight_kg + (1|sample_age) + offset(log(no_years)),
              data=dd_pan_pac,
              family=nbinom2,
              REML=F)


#compare, difference not statistically significant
anova(m1_pac_full,m1_pac_drop)

```

### Estimated marginal means

```{r}

#predict for 1 kg over 100 yr
yr_predict <-100
offset_value_predict <- log(yr_predict)


#Pacific
ef_pac <- effect("age_group", model_selected_pac,
                         given.values=c(log_sed_weight_kg=0), #1kg
                         offset=offset_value_predict)
                         
summary(ef_pac)


#Caribbean
ef_carib <- effect("age_group", model_selected_carib,
                         given.values=c(log_sed_weight_kg=0), #1kg
                         offset=offset_value_predict)

summary(ef_carib)


#combine and add region info for plotting
ef_pac_df <- as.data.frame(ef_pac) %>% 
  add_column(region='Gulf of Panama',.before="age_group")
ef_carib_df <- as.data.frame(ef_carib)%>% 
  add_column(region='Bocas del Toro',.before="age_group")

ef_pan_df <- rbind(ef_carib_df,ef_pac_df) %>% 
  mutate(age_group = fct_relevel(age_group,c("pre_exp","modern")))


#test plot
ggplot(ef_pan_df, aes(x=region, y=fit,color=age_group)) +
  geom_pointrange(aes(ymin=fit-se, ymax=fit+se), position=position_jitterdodge(seed=123),size=1) + 
  theme_bw()+
  labs(y="Denticle accumulation per kg sediment per century",x="Region",color="Age group") +
  scale_x_discrete(limits=c("Gulf of Panama","Bocas del Toro"),labels=c("Gulf of Panama","Bocas del Toro"))+
  scale_y_continuous(trans='log10')+
  theme(aspect.ratio=1)

```

Model-derived magnitude of change estimates.

```{r}

#Caribbean change over time
modern_carib <- ef_pan_df[2,3]
holocene_carib <- ef_pan_df[1,3]

(modern_carib-holocene_carib)/holocene_carib*100 

#Pacific change over time
modern_pac <- ef_pan_df[4,3]
holocene_pac <- ef_pan_df[3,3]

(modern_pac-holocene_pac)/holocene_pac*100 

#Cross-isthmian difference
holocene_pac/holocene_carib 
modern_pac/modern_carib 

```

Emmeans, for each regional model.

```{r}

#Pacific
emm_pacific <- emmeans(
  model_selected_pac,
  ~ age_group)

pairs(emm_pacific)
pairs(emm_pacific, type="response")
confint(pairs(emm_pacific), type="response")
summary(emm_pacific, type="response")


#Caribbean
emm_carib <- emmeans(
  model_selected_carib,
  ~ age_group)

pairs(emm_carib)
pairs(emm_carib, type="response")
confint(pairs(emm_carib), type="response")
summary(emm_carib, type="response")

```

Predictions (for each unique sample_age), for plotting.

```{r}

#Caribbean

#set up data frame to predict
predictions_carib = 
  data.frame(
    basin = dd_pan_carib$basin,
    region = dd_pan_carib$region,
    site_name = dd_pan_carib$site_name,
    sample = dd_pan_carib$sample,
    age_group = dd_pan_carib$age_group,
    sample_age = dd_pan_carib$sample_age,
    unique = dd_pan_carib$unique,
    no_years = yr_predict, #100 yr
    log_sed_weight_kg = 0 #1kg
  ) 

#make predictions
predictions_carib_output <- predict(model_selected_carib, newdata = predictions_carib, type = 'response',se.fit=T)

predictions_carib <- predictions_carib %>% 
  mutate(prediction_fit = predictions_carib_output$fit,
         prediction_se = predictions_carib_output$se.fit) 



#Pacific

#set up data frame to predict
predictions_pac = 
  data.frame(
    basin = dd_pan_pac$basin,
    region = dd_pan_pac$region,
    site_name = dd_pan_pac$site_name,
    sample = dd_pan_pac$sample,
    age_group = dd_pan_pac$age_group,
    sample_age = dd_pan_pac$sample_age,
    unique = dd_pan_pac$unique,
    no_years = yr_predict, #100 yr
    log_sed_weight_kg = 0 #1kg
  ) 

#make predictions
predictions_pac_output <- predict(model_selected_pac, newdata = predictions_pac, type = 'response',se.fit=T)

predictions_pac <- predictions_pac %>% 
  mutate(prediction_fit = predictions_pac_output$fit,
         prediction_se = predictions_pac_output$se.fit) 



#Combine
predictions_combined <- rbind(predictions_carib,predictions_pac) %>% 
  mutate(age_group = fct_relevel(age_group,c("pre_exp","modern")))


#Pulling out the distinct sample_age predictions
predictions_combined_distinct <- predictions_combined %>% 
  select(basin,region,sample,site_name,age_group,sample_age,prediction_fit,prediction_se) %>% 
  distinct()

```

### Model-estimated denticle accumulation rates (Figure 1)

```{r fig.height=3}

fig1 <- ggplot(ef_pan_df, aes(x=region, y=fit,shape=age_group)) +
  geom_linerange(data=predictions_combined_distinct, aes(y=prediction_fit,ymin=prediction_fit-prediction_se,ymax=prediction_fit+prediction_se),position=position_jitterdodge(jitter.width=0.5,seed=10),linetype="solid",alpha=0.4,color="gray70")+
  geom_point(data=predictions_combined_distinct,aes(x=region, y=prediction_fit),alpha=0.4,size=2,position=position_jitterdodge(jitter.width=0.5,seed=10),color="gray70",fill="gray70")+
  geom_pointrange(aes(ymin=fit-se, ymax=fit+se,shape=age_group,fill=region),color='black',fatten=4,size=1.5, position=position_jitterdodge(seed=10)) + 
  theme_bw(base_size=12)+
  labs(y="Denticle accumulation per kg sediment per century",x="",shape="Time period",fill="Region") +
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Sub-recent"))+
  scale_x_discrete(limits=c("Bocas del Toro","Gulf of Panama"),labels=c("Caribbean\n(Bocas del Toro)","Pacific\n(Gulf of Panama)"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  theme(panel.grid.minor = element_blank())+
  theme(legend.position = "bottom")+
  annotation_logticks(sides="l")+
  scale_y_continuous(trans='log10')+
  guides(fill="none")+
  theme(aspect.ratio=1.5)
fig1

```

Save Fig. 1 output.

```{r fig.height=3}

#ggsave("figures/fig1.pdf",plot=fig1,device="pdf",useDingbats=FALSE)

```

### Parametric bootstrap (figure S10)

Caribbean

```{r}

#fitted model
model_selected_carib

#create an object to store the results
boot_change_carib <- rep(NA,500)

#simulate data from the model
sim_carib <- simulate(model_selected_carib,nsim=500)


#refit model to new data
for(i in 1:500){
  #adding simulated data to the df with the empirical observations
  dd_pan_carib$sim_col <- sim_carib[,i] 
  #model with sim data as response, no random effects
  m_boot1 <- glmmTMB(sim_col ~ age_group + log_sed_weight_kg + offset(log(no_years)), 
              data=dd_pan_carib,
              family=nbinom2,
              REML=T)
  #pull coeff for fixed effects from the model run
  fix_effects <- fixef(m_boot1) 
  modern <- exp(fix_effects$cond[1]+fix_effects$cond[2])
  holocene <- exp(fix_effects$cond[1])
  #percent change, using fixed effects estimates from the model
  boot_change_carib[i] <- ((modern-holocene)/holocene)*100 
}

#check distribution
hist(boot_change_carib)


#mean + SD
mean(boot_change_carib) 
sd(boot_change_carib)

```

Pacific

```{r}

#fitted model
model_selected_pac


#create an object to store the results
boot_change_pac <- rep(NA,500)

#simulate data from the model
sim_pac <- simulate(model_selected_pac,nsim=500)


#refit model to new data
for(i in 1:500){
  dd_pan_pac$sim_col <- sim_pac[,i] 
  m_boot1 <- glmmTMB(sim_col ~ age_group + log_sed_weight_kg + offset(log(no_years)), 
              data=dd_pan_pac,
              family=nbinom2,
              REML=T) 
  fix_effects <- fixef(m_boot1) 
  modern <- exp(fix_effects$cond[1]+fix_effects$cond[2])
  holocene <- exp(fix_effects$cond[1])
  boot_change_pac[i] <- ((modern-holocene)/holocene)*100 
}

#check distribution
hist(boot_change_pac)

#central tendency
mean(boot_change_pac)
sd(boot_change_pac)

```

Combine and plot.

```{r fig.height=3}
#combine bootstrap runs
boot_change_pac <- as.data.frame(boot_change_pac) %>% 
  mutate(region="Pacific") %>% 
  rename(rel_change=boot_change_pac)

boot_change_carib <- as.data.frame(boot_change_carib) %>% 
  mutate(region="Caribbean") %>% 
  rename(rel_change=boot_change_carib)

boot_combined <- rbind(boot_change_pac,boot_change_carib)


#plot
figS10 <- ggplot(boot_combined,aes(x=region, y=rel_change, fill=region))+
  geom_hline(yintercept=0,color="gray30",linetype=1)+
  geom_violin(adjust=2,color='gray90',fill="gray90",scale="width",alpha=0.7)+
  geom_boxplot(aes(fill=region),alpha=0.9,outlier.shape = NA,width=0.15)+
  theme_bw(base_size=12)+
  labs(y="Bootstrapped percent change \nbetween pre-exploitation and sub-recent periods",x="Region") +
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_x_discrete(limits=c("Caribbean","Pacific"),labels=c("Caribbean\n(Bocas del Toro)","Pacific\n(Gulf of Panama)"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  coord_cartesian(ylim = c(-100,750))+
  theme(legend.position = "none")+
  theme(aspect.ratio=1.5)
figS10
```

Save output (fig. S10).

```{r fig.height=3}

#ggsave("figures/figS10.pdf",plot=figS10,device="pdf",useDingbats=FALSE)

```

## Supplementary figures

### figure S2

```{r fig.height=5,fig.width=7}

#Add dates
dates_all <- read.csv("data/dates_pacific_panama.csv",header=T,na.strings=".")

#Filter to sites and cores used in this study
bchron_age_depth_LP <- bchron_age_depth %>% 
  filter(region=='Gulf of Panama') %>% 
  filter(!sample %in% c("AT19-1-4","BF18-1-14"))

dates_LP <- dates_all %>% 
  filter(region=='Gulf of Panama') %>%
  filter(!sample %in% c("AT19-1-4","BF18-1-14"))

#Draw cut-off lines
vlines <- data.frame(
  date_locality = c("Contadora", "Saboga", "Isla Iguana"),
  vline1 = c(-2000, -1000, -2000),
  vline2 = c(1900, 1900, 1900)
)

#Plot age CE/BCE
figS2 <-  ggplot(bchron_age_depth_LP,aes(x=predicted_age_median,y=depth))+
  geom_vline(data = vlines, aes(xintercept = vline1),
             color = "gray40", linetype = "dashed") +
  geom_vline(data = vlines, aes(xintercept = vline2),
             color = "gray40", linetype = "dashed") +
  geom_line(aes(color=sample),size=1)+
  geom_ribbon(aes(xmin=(1950-min),xmax=(1950-max),fill=sample),alpha=0.3)+
  geom_point(data=dates_LP,aes(x=U_Th_age,y=depth,color=sample),size=2)+
  labs(x='Age (CE/BCE)', y='Depth (cm)',color='Core',fill='Core')+
  scale_y_reverse()+
  scale_x_continuous(breaks = seq(-6000,2000,1000))+
  theme_bw()+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans"))+
  facet_wrap(~date_locality,nrow=3,ncol=1)+
  theme(aspect.ratio = 0.7)+
  theme(strip.background = element_rect(fill = NA, color = NA),
        strip.text = element_text(size = 16))
figS2

```

Save output (fig. S2).

```{r fig.height=7,fig.width=7}

#ggsave("figures/figS2.pdf",plot=figS2,device="pdf",useDingbats=FALSE,height=12,width=12)

```

### figure S3

```{r}

figS3 <- dd_pan_age_clean %>% 
  filter(basin=="Pacific") %>%
ggplot(aes(x=age_AD_BC, y=age_group))+
  geom_density_ridges(color="deepskyblue3",fill="deepskyblue3",alpha=0.2,size=1) +
  theme_ridges() +
  #Caribbean site ages
  geom_rect(xmin=1651,xmax=2014,ymin=1.7,ymax=1.9,fill='darkgoldenrod2',alpha=0.05)+
  geom_rect(xmin=-5200,xmax=-3700,ymin=0.7,ymax=0.9,fill='darkgoldenrod2',alpha=0.05)+
  theme_bw()+
  labs(x="Sample age (CE/BCE)",y="Time period") +
  scale_y_discrete(limits=c("pre_exp","modern"),labels=c("Pre-exploitation","Sub-recent"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  theme(aspect.ratio=0.5)
figS3

```

Save output (fig. S3).

```{r}

#ggsave("figures/figS3.pdf",plot=figS3,device="pdf",useDingbats=FALSE)
#ggsave("figures/figS3.png",plot=figS3,device="png")

```

### figure S4

Temporal resolution of data.

```{r fig.width=4}

#by region & time period
region_time_means <- dd_pan_age_clean %>% 
  group_by(region,age_group) %>% 
  summarize(mean_years = mean(no_years,na.rm=T),
            sd_years = sd(no_years,na.rm=T),
            se_years = sd(no_years,na.rm=T)/sqrt(length(no_years)))
region_time_means


figS4A <- ggplot(region_time_means,aes(x=region, y=mean_years, shape=age_group))+
  geom_pointrange(aes(ymin=mean_years-se_years, ymax=mean_years+se_years,shape=age_group,fill=region),color='black',fatten=3.5,size=1.5, position=position_jitterdodge(seed=10)) +
  theme_bw()+
  labs(y="Number of years per sample",x="Region",fill="Region",shape="Time period") +
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Sub-recent"))+
  scale_x_discrete(limits=c("Bocas del Toro","Gulf of Panama"),labels=c("Caribbean\n(Bocas del Toro)","Pacific\n(Gulf of Panama)"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  guides(fill="none")+
  scale_y_continuous(trans='log10',limits = c(10, 500))+
  annotation_logticks(sides="l")+
  theme(legend.position = "none")+
  theme(aspect.ratio=1.5)
figS4A


#by site
site_time_means <- dd_pan_age_clean %>% 
  group_by(region,site_name,age_group) %>% 
  summarize(mean_years = mean(no_years,na.rm=T),
            sd_years = sd(no_years,na.rm=T),
            se_years = sd(no_years,na.rm=T)/sqrt(length(no_years)))
site_time_means

figS4B <- ggplot(site_time_means,aes(x=site_name, y=mean_years, shape=age_group))+
  geom_pointrange(aes(ymin=mean_years-se_years, ymax=mean_years+se_years,shape=age_group,fill=region),color='black',fatten=3.5,size=1.5, position=position_jitterdodge(seed=10)) +
  theme_bw()+
  labs(y="",x="Site",fill="Region",shape="Time period") +
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Sub-recent"))+
  scale_x_discrete(limits=c("Sweet Bocas","Casa Blanca","Cayo Adriana","Punta Donato","Isla Iguana","Saboga","Contadora"),labels=c("Sweet\nBocas","Casa\nBlanca","Cayo\nAdriana","Punta\nDonato","Isla\nIguana","Saboga","Contadora"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  guides(fill="none")+
  scale_y_continuous(trans='log10',limits = c(10, 500))+
  annotation_logticks(sides="l")+
  theme(legend.position = "bottom")+
  theme(aspect.ratio=1)
figS4B

```

Combine and save output (fig. S4).

```{r fig.width=5}

figS4 <- figS4A+figS4B+
  plot_annotation(tag_levels = 'A')&
  theme(
    plot.tag = element_text(size = 18, face = "bold"))
figS4


#save output
#ggsave("figures/figS4.png",plot=figS4,device="png")

```

### figure S5

Calculate means for plotting.

```{r}

denticle_means <- dd_pan_age_clean %>% 
  mutate(dd_accum_century = (total_dd_count/sed_weight_kg/no_years)*100) %>% 
  group_by(region) %>% 
  summarize(mean_count = mean(total_dd_count,na.rm=T),
            mean_abundance = mean(dd_kg,na.rm=T),
            mean_accumulation = mean(dd_accum_century,na.rm=T),
            sd_count = sd(total_dd_count,na.rm=T),
            sd_abundance = sd(dd_kg,na.rm=T),
            sd_accumulation = sd(dd_accum_century,na.rm=T))

denticle_means
```

Denticle counts (fig. S5A)

```{r fig.height=3}

figS5A <- ggplot(dd_pan_age_clean, aes(x=total_dd_count, y=site_name, shape=age_group,fill=region)) +
  geom_vline(xintercept=denticle_means$mean_count[2],color="deepskyblue3",linetype=2,linewidth=0.8)+
  geom_vline(xintercept=denticle_means$mean_count[1],color="darkgoldenrod2",linetype=2,linewidth=0.8)+
  geom_point(alpha=0.4,size=2.5,position=position_jitterdodge(jitter.width=0.5,dodge.width=0.7,seed=10),color="gray70",fill="gray70")+
  geom_boxplot(aes(fill=region),alpha=0.8,outlier.shape = NA,width=0.2,position=position_dodge(width = 0.7))+
  theme_bw()+
  labs(x="Number of denticles",y="Site",shape="Time period",fill="Region") +
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Sub-recent"))+
  scale_y_discrete(limits=c("Contadora","Saboga","Isla Iguana","Sweet Bocas","Punta Donato","Cayo Adriana","Casa Blanca"),labels=c("Contadora","Saboga","Isla Iguana","Sweet Bocas","Punta Donato","Cayo Adriana","Casa Blanca"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  theme(panel.grid.minor = element_blank())+
  theme(legend.position = "none")+
  annotation_logticks(sides="b")+
  scale_x_continuous(trans='log10')+
  theme(aspect.ratio=1.75)
figS5A

```

Denticle abundances (fig. S5B)

```{r fig.height=3}

figS5B <- ggplot(dd_pan_age_clean, aes(x=dd_kg, y=site_name, shape=age_group,fill=region)) +
  geom_vline(xintercept=denticle_means$mean_abundance[2],color="deepskyblue3",linetype=2,linewidth=0.8)+
  geom_vline(xintercept=denticle_means$mean_abundance[1],color="darkgoldenrod2",linetype=2,linewidth=0.8)+
  geom_point(alpha=0.4,size=2.5,position=position_jitterdodge(jitter.width=0.5,dodge.width=0.7,seed=10),color="gray70",fill="gray70")+
  geom_boxplot(aes(fill=region),alpha=0.8,outlier.shape = NA,width=0.2,position=position_dodge(width = 0.7))+
  theme_bw()+
  labs(x="Denticles per kg sediment",y="",shape="Time period",fill="Region") +
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Sub-recent"))+
  scale_y_discrete(limits=c("Contadora","Saboga","Isla Iguana","Sweet Bocas","Punta Donato","Cayo Adriana","Casa Blanca"),labels=c("Contadora","Saboga","Isla Iguana","Sweet Bocas","Punta Donato","Cayo Adriana","Casa Blanca"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  theme(panel.grid.minor = element_blank())+
  theme(legend.position = "none")+
  annotation_logticks(sides="b")+
  scale_x_continuous(trans='log10')+
  theme(aspect.ratio=1.75)
figS5B

```

Denticle accumulations (fig. S5C)

```{r fig.height=3}

figS5C <- ggplot(dd_pan_age_clean, aes(x=(total_dd_count/sed_weight_kg/no_years)*100, y=site_name, shape=age_group,fill=region)) +
  geom_vline(xintercept=denticle_means$mean_accumulation[2],color="deepskyblue3",linetype=2,linewidth=0.8)+
  geom_vline(xintercept=denticle_means$mean_accumulation[1],color="darkgoldenrod2",linetype=2,linewidth=0.8)+
  geom_point(alpha=0.4,size=2.5,position=position_jitterdodge(jitter.width=0.5,dodge.width=0.7,seed=10),color="gray70",fill="gray70")+
  geom_boxplot(aes(fill=region),alpha=0.8,outlier.shape = NA,width=0.2,position=position_dodge(width = 0.7))+
  theme_bw()+
  labs(x="Denticles per kg sediment\nper century",y="",shape="Time period",fill="Region") +
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Sub-recent"))+
  scale_y_discrete(limits=c("Contadora","Saboga","Isla Iguana","Sweet Bocas","Punta Donato","Cayo Adriana","Casa Blanca"),labels=c("Contadora","Saboga","Isla Iguana","Sweet Bocas","Punta Donato","Cayo Adriana","Casa Blanca"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  theme(panel.grid.minor = element_blank())+
  theme(legend.position = "right")+ 
  annotation_logticks(sides="b")+
  scale_x_continuous(trans='log10')+
  theme(aspect.ratio=1.75)
figS5C

```

Combine and save output (fig. S5).

```{r fig.height=3,fig.width=8,warning=F}

figS5 <- figS5A+figS5B+figS5C+
  plot_annotation(tag_levels = 'A')&
  theme(
    plot.tag = element_text(size = 18, face = "bold"))
figS5


#save output
#ggsave("figures/figS5.pdf",plot=figS5,device="pdf",useDingbats=FALSE)

```

### figure S7

Model predictions vs. raw data, with time uncertainty visualized.

```{r fig.height=3}

figS7 <- ggplot(ef_pan_df, aes(x=region, y=fit,shape=age_group)) +
  geom_linerange(data=dd_pan_age_clean,aes(y=((total_dd_count/sed_weight_kg/no_years)*100),ymin=((total_dd_count/sed_weight_kg/no_years_75)*100),ymax=((total_dd_count/sed_weight_kg/no_years_25)*100)),position=position_jitterdodge(jitter.width=0.6,seed=10),linetype="solid",alpha=0.4,color="gray80")+
  geom_point(data=dd_pan_age_clean,aes(x=region, y=(total_dd_count/sed_weight_kg/no_years)*100),alpha=0.4,size=2,position=position_jitterdodge(jitter.width=0.6,seed=10),color="gray80",fill="gray80")+
  geom_pointrange(aes(ymin=fit-se, ymax=fit+se,shape=age_group,fill=region),color='black',fatten=4,size=1.4, position=position_jitterdodge(seed=10)) + 
  theme_bw(base_size=12)+
  labs(y="Denticle accumulation per kg sediment per century",x="Region",shape="Time period",fill="Region") +
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Sub-recent"))+
  scale_x_discrete(limits=c("Bocas del Toro","Gulf of Panama"),labels=c("Caribbean\n(Bocas del Toro)","Pacific\n(Gulf of Panama)"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  theme(panel.grid.minor = element_blank())+
  theme(legend.position = "bottom")+
  annotation_logticks(sides="l")+
  scale_y_continuous(trans='log10')+
  guides(fill="none")+
  theme(aspect.ratio=1.5)
figS7

```

Save output (fig. S7).

```{r}

#ggsave("figures/figS7.png",plot=figS7,device="png")

```

### figure S14

Variation in sample weights.

```{r fig.width=4}

figS14 <- ggplot(dd_pan_age_clean,aes(x=region, y=sed_weight_kg, fill=region, shape=age_group))+
  geom_point(alpha=0.4,size=2,position=position_jitterdodge(jitter.width=0.5,dodge.width=0.7,seed=10),color="gray70",fill="gray70")+
  geom_boxplot(aes(fill=region),alpha=0.8,outlier.shape = NA,width=0.2,position=position_dodge(width = 0.7))+
  theme_bw()+
  labs(y="Sample weight (kg)",x="Region",shape="Time period",fill="Region") +
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Sub-recent"))+
  scale_x_discrete(limits=c("Bocas del Toro","Gulf of Panama"),labels=c("Caribbean\n(Bocas del Toro)","Pacific\n(Gulf of Panama)"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  guides(fill="none")+
  annotation_logticks(sides="l")+
  scale_y_continuous(trans='log10')+
  theme(legend.position = "bottom")+
  theme(aspect.ratio=1.5)
figS14

```

Save output (fig. S14).

```{r fig.width=4}

#ggsave("figures/figS14.pdf",plot=figS14,device="pdf",useDingbats=FALSE)

```

### figure S15

```{r}

#denticle counts x sediment weight
figS15A <- ggplot(dd_pan_age_clean,aes(x=sed_weight_kg,y=total_dd_count))+
  geom_point(aes(shape=age_group,fill=region),color='gray30',alpha=0.8,size=2.5)+
  theme_bw()+
  labs(x="Sample weight (kg)",y="Number of denticles",shape="Time period",fill="Region") +
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Sub-recent"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  guides(fill="none")+
  theme(legend.position = "none")+
  scale_y_continuous(trans='log10')+
  scale_x_continuous(trans='log10')+
  theme(aspect.ratio=1)+
  facet_grid(~basin,scales = "free")+
  theme(strip.background = element_rect(fill = NA, color = NA),
        strip.text = element_text(size = 16))
figS15A


#sediment weight x no_years
figS15B <- ggplot(dd_pan_age_clean,aes(x=sed_weight_kg,y=no_years))+
  geom_point(aes(shape=age_group,fill=region),color='gray30',alpha=0.8,size=2.5)+
  theme_bw()+
  labs(x="Sample weight (kg)",y="Number of years per sample",fill="Region",shape="Time period") +
  scale_fill_manual(values=c("darkgoldenrod2","deepskyblue3"))+
  scale_shape_manual(values=c(21,24),labels=c("Pre-exploitation","Sub-recent"))+
  theme(axis.text = element_text(size=12),axis.title= element_text(size=16))+
  theme(legend.text = element_text(size=12),legend.title = element_text(size=16))+
  theme(axis.text=element_text(color="black",family="sans"),axis.title= element_text(color="black",family="sans",hjust=0.5))+
  guides(fill="none")+
  theme(legend.position = "bottom")+
  scale_y_continuous(trans='log10')+
  scale_x_continuous(trans='log10')+
  theme(aspect.ratio=1)+
  facet_grid(~basin,scales = "free")+
  theme(strip.background = element_rect(fill = NA, color = NA),
        strip.text = element_text(size = 16))
figS15B

```

Combine and save output (fig. S15).

```{r fig.width=4,fig.height=4,warning=F}

figS15 <- figS15A/figS15B+
  plot_layout(widths = c(3,-8,3))+
  plot_annotation(tag_levels = 'A')&
  theme(
    plot.tag = element_text(size = 18, face = "bold"))
figS15


#save output
#ggsave("figures/figS15.png",plot=figS15,device="png")

```
